---
title: "Howard County Council Supersizing, Part 1"
author: "Frank Hecker"
date: "`r format(Sys.time(), '%Y-%m-%d')`"
output:
  html_document:
    fig_width: 8
    fig_height: 5
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

In this document I create a shapefile to be used for Howard County Council
redistricting calculations. The shapefile combines precinct boundaries,
precinct populations (including breakdowns by race), and election results.

For those readers unfamiliar with the [R statistical software][r] and
the additional [Tidyverse software][tv] I use to manipulate and plot
data, I’ve included some additional explanation of various steps.  For
more information check out the various [ways to learn more about the
Tidyverse][learn].

[r]: https://www.r-project.org/
[tv]: https://www.tidyverse.org/
[learn]: https://www.tidyverse.org/learn/

## Setup and data preparation

### Libraries

I use the following packages for the following purposes:

* tidyverse: do general data manipulation.
* readxl: read Microsoft Excel spreadsheets.
* sf: manipulate geospatial data.
* tools. compute MD5 checksums.

```{r libraries, message=FALSE, warning=FALSE}
library(tidyverse)
library(readxl)
library(sf)
library(tools)
```

### Data sources

I use data from the following sources; see the References section
below for more information:

* Boundaries and populations for Howard County precincts are from the Maryland
  Department of Planning data used for Congressional and Maryland legislative
  redistricting. I use the data from the 2010 census since the data for 2020
  are not yet available.
* Election results for Howard County precincts are from the Maryland Board of
  Elections. I use the results from the 2010 general election since it
  featured local Howard County races (including for county executive and
  county council) and the precinct number and boundaries match the 2010
  redistricting data from the Maryland Department of Planning.

### Downloading the data

I download the following files:

* Precinct shapefile from the 2010 Maryland redistricting effort. This
  shapefile also contains unadjusted and adjusted population figures
  broken down by race.
* Howard County 2010 general election results for county races.

```{r download}
download.file("https://planning.maryland.gov/Redistricting/Documents/2010zip/precinct2010.zip",
              "precinct.zip",
              method = "curl")
download.file("https://elections.maryland.gov/elections/2010/election_data/Howard_By_Precinct_2010_General.csv",
              "election_2010.csv",
              method = "curl")
```
I check the MD5 value for the files, and stop if the contents are not what
are expected.

```{r md5_check}
stopifnot(md5sum("precinct.zip") == "d620e6d5e26e56052b02da2588e4835a")
stopifnot(md5sum("election_2010.csv") == "bd071b938df5a3c5080c017cb5993675")
```

I unzip the precinct shapefile data.

```{r unzip}
unzip("precinct.zip", overwrite = TRUE)
```

### Reading in and preparing the data

I first read in the shapefile containing adjusted 2010 populations for
all Maryland precincts, and keep only the precinct data and geometry
for Howard County.

For some reason the shapefile population data omits the adjusted population
values for multiracial people, so I create a new variable `ADJ_MUL` calculated
as the difference between the total adjusted population value and the adjusted
population values for the other racial categories.

```{r precinct_sf}
precinct_sf <- st_read("Precinct2010_83NadMeters.shp") %>%
  filter(COUNTY == "24027") %>%
  mutate(ADJ_MULTI = ADJ_POPULA -
           (ADJ_WHITE + ADJ_BLACK + ADJ_ASIAN +
              ADJ_AMINDI + ADJ_HAWAII + ADJ_OTHER))
```

I then read in the 2010 election data for Howard County, and keep only the
Democratic and Republican totals for the county executive race and
the five county council races. (Note that there were competitive races for
all five council districts in 2010.)

In order to prepare the data for merging with the shapefile data, I combine
the fields for the office and for the party into a single field, get rid of
all the unneeded fields, and then use `pivot_wider()` to convert the different
rows containing vote totals for the county executive and county council races
into different columns.

```{r hc_election}
election_2010 <- read_csv("election_2010.csv") %>%
  filter(`Office Name` == "County Executive" |
           `Office Name` == "County Council") %>%
  filter(Party == "DEM" | Party == "REP") %>%
  mutate(VTD = paste("24027",
                     substr(`Election District`, 2, 3),
                     "-",
                     `Election Precinct`,
                     sep = "")) %>%
  mutate(`Office by Party` = paste(`Office Name`, "-", `Party`, sep = "")) %>%
  select(VTD, `Office by Party`, `Election Night Votes`) %>%
  pivot_wider(names_from = `Office by Party`,
              values_from = `Election Night Votes`,
              values_fill = 0)
```

I join the two tables `precinct_sf` and `election_2010` using their
common field `VTD` to create a new shapefile containing precinct boundaries,
(adjusted) populations (broken down by race), and 2010 election results for
county executive and county council.

```{r redistricting_sf}
redistricting_sf <- inner_join(precinct_sf, election_2010, by = "VTD")
```

Finally, I write the shapefile out to disk, to be used as input to the
AutoRedistrict application.

```{r st_write}
st_write(redistricting_sf, "redistricting.shp", append = FALSE)
```

## Appendix

### Caveats

The election data omits votes cast by third parties or for write-in candidates.
In practice the number of these votes is very small, and should not affect the
redistricting results.

### References

Data derived from the 2010 census and used for Maryland redistricting is
available from the "[Precinct Population Count and Maps][ppc]" page
published by the Maryland Department of Planning. The data is combined with
precinct boundaries to create the [2010 precinct shapefile][ps] (distributed
as a zip file).

[ppc]: https://planning.maryland.gov/Redistricting/Pages/2010/precinct.aspx
[ps]: https://planning.maryland.gov/Redistricting/Documents/2010zip/precinct2010.zip

Precinct-level results for the Howard County 2010 general election are
available from the "[Data Files for the 2010 Gubernatorial Election
Results][df2010]" page published by the Maryland Board of Elections. This
analysis uses the [precinct results for Howard County][prhc].

[df2010]: https://elections.maryland.gov/elections/2010/election_data/index.html
[prhc]: https://elections.maryland.gov/elections/2010/election_data/Howard_By_Precinct_2010_General.csv

### Suggestions for future work

This analysis needs to be updated as soon as the data from the 2020 census is
released by Maryland Department of Planning.

### Environment

I used the following R environment in doing the analysis above:

```{r}
sessionInfo()
```

### Source code

You can find the source code for this analysis and others at my
[hocodata public code repository][hd].  This document and its source
code are available for unrestricted use, distribution and modification
under the terms of the [Creative Commons CC0 1.0 Universal (CC0 1.0)
Public Domain Dedication][cc0].  Stated more simply, you’re free to do
whatever you’d like with it.

[hd]: https://gitlab.com/frankhecker/hocodata
[cc0]: https://creativecommons.org/publicdomain/zero/1.0/
