---
title: "Howard County 2014 Residential Building Permits, Example 3"
author: "Frank Hecker"
date: "February 16, 2015"
output: html_document
---

## Introduction

This is a continuation of my previous examples of analyzing building
permit data from the `[data.howardcountymd.gov][data]` site to answer the
following question: Which localities within Howard County, Maryland,
saw the most residential building permits issued in 2014?

[data]: http://data.howardcountymd.gov/

In this version I create a map of Howard County with the varying
number of permits shown using variations in color from one zip
code area to another.

[data.howardcountymd.gov]: http://data.howardcountymd.gov/

## Load packages

For this analysis I’ll again be using the [R statistical
package][] run from the [RStudio][] development environment, along
with the [dplyr package][] to do data manipulation and the [ggplot2
package][] to draw the map.

[R statistical package]: http://r-project.org
[RStudio]: http://www.rstudio.com/products/rstudio
[dplyr package]: http://cran.rstudio.com/web/packages/dplyr/vignettes/introduction.html
[ggplot2 package]: http://ggplot2.org

```{r}
library("dplyr", warn.conflicts = FALSE)
library("ggplot2")
```

I also need to load several R packages used to manipulate spatial
data in R. I first load the [sp package][], a prerequisite for using
the other packages. I use the [rgdal package][] to load spatial data
on zip code boundaries downloaded from `data.howardcountymd.gov`,
and the [rgeos][] and [maptools][] packages to manipulate the spatial
data once loaded. The maptools package checks on the presence of
the rgeos package, so I load rgeos first and then maptools.

[sp package]: http://cran.r-project.org/web/packages/sp/index.html
[rgdal package]: http://cran.r-project.org/web/packages/rgdal/index.html
[rgeos]: http://cran.r-project.org/web/packages/rgeos/index.html
[maptools]: http://cran.r-project.org/web/packages/maptools/index.html

```{r}
library("sp")
library("rgdal")
library("rgeos")
library("maptools")
```

I should note here (having forgotten to do it in the previous
examples) that none of the packages above are part of the base R
system, and so I had to install them using the
`install.packages()` function. The mapping packages also require
installing sets of underlying mapping libraries. Doing this is not
trivial, and varies from operating system to operating system.
Unfortunately I don’t have space in this document to describe the
entire process.

## Loading the data

First I download the CVS-format data relating to issuance of building
permits from the `data.howardcountymd.gov` site and store it in a
local file `hoco-building-permits.csv`. (This is the same process as
in the previous examples.)

```{r}
download.file("https://data.howardcountymd.gov/geoserver/ows?service=WFS&version=1.0.0&request=GetFeature&typeName=general:Permits_View_Building_New&outputFormat=csv",
              "hoco-building-permits.csv", method = "curl")
```

Then I download spatial data from `data.howardcountymd.gov`
spcifying the boundaries of Howard County zip code areas. The
Howard County GIS Division rather nicely offers this data in
a variety of formats; I chose to use the GeoJSON format, and 
thus stored the data locally in a file `zipcodes.json`.

```{r}
download.file("https://data.howardcountymd.gov/geoserver/ows?service=WFS&version=1.0.0&request=GetFeature&typeName=general:Zipcodes&outputFormat=application/json", "zipcodes.json", method = "curl")
```

Next I read the CSV data for building permits and convert it into a
data frame. (Again, this is the same process as in the last two
examples.)

```{r}
permits <- read.csv("hoco-building-permits.csv", stringsAsFactors = FALSE)
```

Finally I read in the GeoJSON zip code data. The `readOGR()` function
is part of the rgdal package; since the function understands multiple
format I have to tell it which format this file is in.

```{r}
zips <- readOGR("zipcodes.json", "OGRGeoJSON")
str(zips@data)
```

The `zips` variable is not a regular data frame but rather is a
special type of data structure, a “SpatialPolygonsDataFrame”
containing both a regular data frame (in `zips@data`) and a list
of 25 “polygons” containing map data for the 25 Howard County
zip codes.

## Data processing

I first need to repeat the processing steps from the prior examples to
get a count of residential permits issued for each zip code in
2014. The main difference in this example is that I choose to work
only with zip codes and do not need the locality names.

```{r}
permits_by_zip <- permits %>%
    select(Issued_Date, Permit_Type_2, Detailed_Permit_Type, City, Zip) %>%
    filter(Permit_Type_2 == "Residential") %>%
    filter(grepl("/2014$", Issued_Date)) %>%
    group_by(Zip) %>%
    summarise(Permits = n()) %>%
    arrange(desc(Permits))
```

Now I have to further modify the `permits_by_zip` data frame to add
data for zip codes for which there were no permits issued in 2014. (I
neglected to do this the first time, and the resulting map was messed
up something awful.)

The `permits_by_zip` data frame includes the following zip codes:

```{r}
sort(permits_by_zip$Zip)
```

The `zips@data` data frame within the `zips` data structure contains
the following zip codes:

```{r}
sort(unique(zips@data$ZIPCODE))
```

I want to find all zip codes that are in the zip code map but not in
`permits_by_zip`. I first convert the `ZIPCODE` values from
`zips@data` into integers, to match the `permits_by_zip$Zip` values.
(I have to convert the `ZIPCODE` values first to characters and then
to integers because the values are _factors_, an R data type that has
some odd properties.)


```{r}
map_zips <- as.integer(as.character(zips@data$ZIPCODE))
str(map_zips)
```

I then use the `setdiff()` function to find the set of zip codes in
`map_zips` that are not in `permits_by_zip$Zip`. (I reference the
`setdiff()` function as `base:setdiff` to avoid a name conflict with a
function of the same name in the dplyr package.)  I store the
resulting set of zips codes in a variable named `Zip`, for reasons
which will become apparent in the next step.


```{r}
Zip <- base::setdiff(map_zips, permits_by_zip$Zip)
Zip
```

Next I use the variable `Zip` to create a data frame `permits_by_zip2`
with a single variable `permits_by_zip2$Zip` as the first and only
column. Then I add a new column `permits_by_zip2$Permits` with all
zero values. Finally I combine the original `permits_by_zip` data
frame with the newly-created `permits_by_zip2` data frame to create a
modified `permits_by_zip` data frame with rows for the formerly
missing zip codes. (This is why I chose the name `Zip` above for the
column in `permits_by_zip2`, to match the corresponding column name in
`permits_by_zip`.)

Since I might want to reuse this code for future years, I allow for
the possibility that there are no zip codes without permits issued in
a given year, and do all this only if the set of missing zip codes is
non-empty.

```{r}
if (length(Zip) > 0) {
    permits_by_zip2 <- as.data.frame(Zip)
    permits_by_zip2$Permits <- 0
    permits_by_zip <- bind_rows(permits_by_zip, permits_by_zip2)
}
```

I then print the entire resulting data frame to show the added rows
for the zip codes with no permits issued in 2014.

```{r}
print.data.frame(permits_by_zip)
```

## Mapping the number of permits by zip code

Now comes the fun part: actually mapping the data. I first use the
`plot()` function in base R to verify that the `zips` data structure
actually contains the necessary data to produce a zip code map of
Howard County.

```{r}
plot(zips)
```

So far so good. I now have two possible strategies I can follow. What
I want to do is to color each of the zip code areas differently
according to the number of permits issued in that particular
area. (Fun fact: This is called a [choropleth map][].)

[choropleth map]: http://en.wikipedia.org/wiki/Choropleth_map

One way to do this is to use the `plot()` function in base R and
provide a vector of colors to use in coloring each zip code area
(polygon), with the colors calculated according to the number of
permits issued. The `plot()` function doesn’t know how to do this
calculation itself, so I would have to write code to do it.

The other approach is to use the `ggplot()` function from the ggplot2
package. The `ggplot()` function already knows how to pick colors based
on data values. However it doesn’t understand the `zips` data
structure (a SpatialPolygonsDataFrame), so I have to convert that data
into the type of data frame that `ggplot()` expects.

In this analysis I follow the second strategy. I got the basic idea
for how to do this from the answers to a Stack Overflow question “[R
ggplot2 merge with shapefile and csv data to fill
polygons][stackoverflow]”. I have to add a new column `id` to the data
frame `zip@data` inside `zips`, use the function `fortify()` to
convert the polygons in `zips` into individual latitude/longitude
pairs grouped by `id`, then merge that data with the data in `zip@data`
to create a new data frame `zips_df`.

[stackoverflow]: http://stackoverflow.com/questions/19791210/r-ggplot2-merge-with-shapefile-and-csv-data-to-fill-polygons

```{r}
zips@data$id <- rownames(zips@data)
zips_points <- fortify(zips, region = "id")
zips_df <- merge(zips_points, zips@data, by = "id")
str(zips_df)
```

The resulting variable `zips_df` looks more like the sort of data
frame we’ve in previous examples; each row (observation) corresponds
to a point in the set of polygons representing the zip code
boundaries, and includes values for the zip code with which the point
is associated.

Now I want to combine the `zips_df` data frame with the
`permits_by_zip` data to create a consolidated data set with the
number of permits included. To do that I first create a new variable
`zips_df$Zip` that is of the same integer data type as the variable
`permits_by_zip$Zip`, and then join the two data frames together based
on that common variable. The `left_join()` function (from the dplyr
package) takes all the rows in the `zips_df` data frame for each zip
code and adds the `Permits` column from the row in `permits_by_zip`
that has a matching zip code.

```{r}
zips_df <- mutate(zips_df, Zip = as.integer(as.character(ZIPCODE)))
zips_df <- left_join(zips_df, permits_by_zip, by = "Zip")
str(zips_df)
```

Now I have something `ggplot()` can deal with. I tell `ggplot()` to
plot `zips_df` using the `long` (longitude) values on the x axis and
the `lat` (latitude) values on the y axis. Then I add to the graph
object `g` the drawing of actual polygons, grouping the points
together that correspond to the various polygons, with the color of
each polygon (the “fill” value) determined based on the value of the
`zips_df$Permits` value for that polygon.

```{r}
g <- ggplot(zips_df, aes(x = long, y = lat, group = group, fill = Permits))
g <- g + geom_polygon(aes(group = group, fill = Permits))
print(g)
```

This map doesn’t look bad as a first attempt. The colors do reflect
the number of permits, and there’s even a nice legend to specify
how they match up. However the map looks somewhat squished in the
horizontal direction, and I don’t really need or want the latitude
and longitude values on the axes.

Those are relatively easy things to fix. However what I really want to
do is more complicated: I'd like to label each area with its zip code
value. To do this I first need to figure out where to put the zip code
labels on the map. I use the `coordinates()` function to find the
centers (more correctly, the [centroids][]) of each of the 25 polygons
in the original `zips` data structures that store the zip code
boundaries.

[centroids]: http://en.wikipedia.org/wiki/Centroid

```{r}
zips_centers = coordinates(zips)
str(zips_centers)
```

The `zips_centers` variable is a matrix of 25 rows and 2 columns. To
use it with `ggplot()` I have to convert it into a data frame,
including giving the two columns their correct names (“long” and
“lat”). I then add a third column containing the zip code values from
the `zips@data` data frame within the `zips` data structure. (This
works because the zip code values in `zips@data` are in the same
order as the polygons corresponding to those zip codes.)

```{r}
zips_centers_df <- as.data.frame(zips_centers)
names(zips_centers_df) <- c("long", "lat")
zips_centers_df$Zip = as.character(zips@data$ZIPCODE)
str(zips_centers_df)
```

Now I have two data frames, `zips_df` and `zips_centers_df`, and I
want to plot them on the same map. To do this I start off with an
“empty” graph object, created by calling `ggplot()` with no
arguments. I next add a layer consisting of polygons for the zip code
areas, with latitude and longitude data taken from `zips_df` and the
polygons colored according to the `Permits` variable in
`zips_df`. Then I add a second layer consisting of text strings for
the zip codes themselves, with latitude and longitude data taken from
`zips_centers_df`, the text labels themselves from the `Zip` variable
in `zips_centers_df`, and the text colored and sized as indicated.  (I
had to experiment a little to find a text size that worked well.)

```{r}
g <- ggplot()
g <- g + geom_polygon(data = zips_df, aes(x = long, y = lat,
                      group = group, fill = Permits))
g <- g + geom_text(data = zips_centers_df, aes(x = long, y = lat, label = Zip),
                   size = 2.5, colour = "white", show_guide = FALSE)
```

Finally I made a few other minor fixes to the graph: Setting the
coordinates so that the map would have a proper aspect ratio, removing
the text strings on the x and y axes, and giving the map a suitable
title.

```{r}
g <- g + coord_equal()
g <- g + theme(axis.title = element_blank(), axis.text = element_blank())
g <- g + ggtitle("Howard County, Maryland, 2014 Residential Building Permits")
print(g)
```

There are a few other things I might want to do in a perfect world,
but as it is the resulting graph looks reasonably professional.

## Conclusion

In this example I made the big leap from tables and bar charts to
actual data-driven maps. It was a fair amount of trouble to install
all the necessary geospatial software (all free, by the way) and
figure out how to use the geospatial data with ggplot2, but I think
the results were well worth it.

That concludes this example of analyzing Howard County building permit
data. Unless I can think of something else worth doing this will be
the last example in this particular series.

## Appendix

I used the following R environment in doing the analysis for this
example:

```{r sessionInfo}
sessionInfo()
```

The underlying GDAL and GEOS libraries for the rgeos and rgdal
packages are from the KyngChaos [GDAL Complete][] distribution version
1.11 for Mac OS X.

[GDAL Complete]: http://www.kyngchaos.com/software/frameworks#gdal_complete

You can find the source code for this analysis and others at my
[HoCoData repository on GitHub][hocodata]. This document and its
source code are available for unrestricted use, distribution and
modification under the terms of the Creative Commons [CC0 1.0
Universal (CC0 1.0) Public Domain Dedication][CC0].  Stated more
simply, you’re free to do whatever you‘d like with it.

[hocodata]: https://github.com/frankhecker/hocodata
[CC0]: https://creativecommons.org/publicdomain/zero/1.0/
