---
title: "Howard County 2014 Residential Building Permits, Example 1"
author: "Frank Hecker"
date: "February 14, 2015"
output: html_document
---

## Introduction

This is a simple example of analyzing building permit data from the
`data.howardcountymd.gov` site. I want to answer the following
question: Which localities within Howard County, Maryland, saw the
most residential building permits issued in 2014?

This question is *not* the same as asking which places saw the most
housing units built, because a single permit could cover multiple
dwellings, for example a multi-unit apartment complex. Nevertheless
the answer will give a rough picture of where residential development
is occurring in the county.

## Load needed libraries

For this analysis I’ll be using the [R statistical
package](http://r-project.org), run from the [R
Studio](http://www.rstudio.com/products/rstudio/) development
environment. To help make the analysis easier I’ll first use the
`library()` function to load an additional R *package*, add-on
software that provides additional capability not in the base R
software.

The [dplyr
package](http://cran.rstudio.com/web/packages/dplyr/vignettes/introduction.html)
provides functions for manipulating datasets in a fairly intuitive way
(at least, more intuitive than is often the case with R). Because the
package provides functions whose names conflict with those elsewhere
in R, I call the `library()` function with the argument
`warn.conflicts = FALSE` to suppress associated warning messages.

```{r initialize}
library("dplyr", warn.conflicts = FALSE)
```

## Loading the data

First I download the CVS-format data relating to issuance of building
permits from the `data.howardcountymd.gov` site and store it in a
local file `hoco-building-permits.csv`.

```{r}
download.file("https://data.howardcountymd.gov/geoserver/ows?service=WFS&version=1.0.0&request=GetFeature&typeName=general:Permits_View_Building_New&outputFormat=csv",
              "hoco-building-permits.csv", method = "curl")
```

Next I read the CSV data and convert it into a *data frame*, a
standard R object for representing datasets. A data frame is similar
to a spreadsheet: It has multiple columns representing different
variables (one per column), and multiple rows representing different
sets of values for those variables (one set per row). One major
difference from a traditional spreadsheet is that in R you can refer
to the variables (columns) using meaningful names, rather than using
letters like ‘A’, ‘B’, and so on.

I use the `stringsAsFactors = FALSE` argument to the `read.csv()`
function to keep character strings in the data as character strings,
and not convert them into *factors* (an R data type used for
*categorical variables*, i.e., variables whose values are like answers
to a multiple-choice question).

```{r}
permits <- read.csv("hoco-building-permits.csv", stringsAsFactors = FALSE)
```

Now let’s take a look at what’s in the dataset; the `str()` function
shows the overall structure of the data frame (hence the function
name):

```{r}
str(permits)
```

The first thing to notice is that there are `r ncol(permits)`
variables (columns) in the data frame, ranging from `FID` to `geom`.

The second thing to notice is that there are a total of
`r nrow(permits)` rows in the dataset (“obs.”, or observations), each
representing a single issued building permit.  Since this dataset gets
continually updated as more permits are issued, as time goes on the
number of rows in the dataset will grow.

Finally, the `str()` function prints the first few values for each of
the variables (corresponding to the first few rows of the data
frame). These values are numbers with decimals (‘num’), character
strings (‘str’), or numbers without decimals (‘int’, for integer).

## Choosing which data to work with

As it turns out I don’t need all the data in the building permits
dataset in order to answer the question I asked. My next step is
therefore to reduce the amount of data I’m working with. In
spreadsheet terms this is analogous to selecting only the columns and
rows I want, and deleting the rest. First I select the variables
(columns) I want, and then I filter the dataset for the observations
(rows) I want.

Some of the variables in the permits dataset are not of immediate
interest, including `FID` and `Permit_ID` (internal identifiers for
each permit), `X`, `Y`, and `geom` (data for mapping), and
`Is_Capital_Project` (irrelevant for this example). I also don’t care
about the entire address, just the city and zip portions.

Looking a bit further, the variable `Permit_Type` appears to always
have the same value, ‘Building’. I can test this using the `unique()`
function to print out all the unique values the variable `Permit_Type`
takes on.  (The syntax `permits$Permit_Type`, with the ‘$’ separating
`permits` and `Permit_Type`, is how I tell R that I am interested in
the variable `Permit_Type` within the data frame `permits`.)

```{r}
unique(permits$Permit_Type)
```

As I thought, this variable always has the same value, presumably
since all the data in this data set is related to building
permits. Since the variable doesn’t add any useful information I can
ignore it as well.

I then start the process of cutting down the data I’m dealing with by
creating a new data frame with just the variables I need. The
`select()` function takes as its first argument a data frame, with the
remaining arguments naming the variables I want to keep. The function
returns a new data frame, which I assign to the variable `permits1`.

```{r}
permits1 <- select(permits, Issued_Date, Permit_Type_2,
                   Detailed_Permit_Type, City, Zip)
str(permits1)
```

The new data frame `permits1` has only `r ncol(permits1)` variables,
but still has the full set of `r nrow(permits1)` observations. Having
picked the variables I want, I now want to cut down the data set to
include only the permits I’m interested in.

My original question was about residential permits, so I want to
filter the data to eliminate non-residential permits. Looking at the
values in `permits2` it appears that the variable `Permit_Type_2` will
allow me to figure out which permits are for residential construction
and which are not. I confirm this by using the `unique()` function to
print out all the unique values that the variable `Permit_Type_2`
takes on.

```{r}
unique(permits1$Permit_Type_2)
```

As I thought, this variable has the value ‘Residential’ for
residential building permits and ‘Commercial’ for non-residential
permits. I can therefore filter the data again to create a new data
frame `permits2` that contains only residential permits. (The operator
‘==’ is a check for equality, that is, that the value of the variable
`Permit_Type_2` is the character string ‘Residential’.)

```{r}
permits2 <- filter(permits1, Permit_Type_2 == "Residential")
str(permits2)
```

I next want to restrict myself to permits issued in 2014, using the
`Issued_Date` variable.

Date handling in R can be somewhat clunky at times, with multiple ways
of handling dates depending on what you want to do with them. In this
case the easiest way to identify 2014 permits is to take `Issued_Date`
as a character string and look for the last part of the string being
‘/2014’, using the `grepl()` function.

(The ‘grep’ part of the name ‘grepl’ comes from the Unix/Linux command
`grep`; for more information see the relevant [Wikipedia
article](http://en.wikipedia.org/wiki/Grep). The ‘l’ part of the name
indicates that the function returns a logical value TRUE or FALSE,
depending on whether the string in question was found or not. The ‘\$’
part of the search string ‘/2014$’ tells the function to look for
‘/2014’ only at the end of the string.)

```{r}
permits3 <- filter(permits2, grepl("/2014$", Issued_Date))
str(permits3)
```

The resulting data frame `permits3` has the same `r ncol(permits2)`
variables as the data frame `permits1`, but has only
`r nrow(permits3)` observations.

## Counting the number of permits for each locality

I now want to count the number of permits issued for each locality
within Howard County. There are two variables that specify the
location for which the permit was issued: `City` and `Zip`. I again
use the `unique()` function to determine what values these variables
contain:

```{r}
unique(permits3$City)
unique(permits3$Zip)
```

It looks as if there are more zip codes than locality names, probably
because both Columbia and Ellicott City contain multiple zip
codes. It’s useful to be able to go down to individual zip codes, but
the zip codes are hard to interpret for people not familiar with the
area.

To get the best of both approaches I use the `mutate()` function to
create a new data frame with the `City` and `Zip` variables
consolidated into a single variable `CityZip`. The `paste()` function
pastes together the values of the two original variables with a slash
character used as a separator.

```{r}
permits4 <- mutate(permits3, CityZip = paste(City, Zip, sep = "/"))
str(permits4)
```

I’m now ready to count the number of residential building permits in
each locality. I first use the function `group_by()` to group the
permits based on the values of the `CityZip` variables, and then use
the function `summarise()` to count the number of permits in each
group.

```{r}
permits5 <- group_by(permits4, CityZip)
permits6 <- summarise(permits5, Permits = n())
str(permits6)
```

The new data frame has only `r nrow(permits6)` rows, one per locality,
and only two fields: `CityZip`, the variable I grouped by, and
`Permits`, the variable containing the number of permits in each group
(returned by the `n()` function).

## Displaying the number of permits by locality

Since I’m interested in which localities had the most permits issued I
sort the rows using the `arrange()` function to put them in descending
order based on the `Permits` variable.

```{r}
permits7 <- arrange(permits6, desc(Permits))
```

Finally I print the entire data frame to show all of the localities
and the number of residential building permits issued for each one in
2014.

```{r}
print.data.frame(permits7)
```

## Conclusion

Given that the two Ellicott City zip codes cover a relatively large
and populous area of Howard County (see the [zip code
map](https://data.howardcountymd.gov/mapgallery/misc/Zipcodes22X34.pdf)
on `data.howardcountymd.gov`), it’s not surprising that Ellicott City
saw the most residential permits issued in 2014. Marriottsville,
Fulton, and Elkridge also saw a lot of residential
development. However it’s interesting to note that East Columbia
(21045 zip code) had only two residential building permits issued in
all of 2014.

That concludes this example of analyzing Howard County building permit
data. In the next example I’ll show how to streamline the analysis and
then create a graph from the summary data.

## Appendix 1: R environment

I used the following R environment in doing the analysis for this
example:

```{r sessionInfo}
sessionInfo()
```

## Appendix 2: Public domain dedication

I am making this work available under the terms of the Creative
Commons [CC0 1.0 Universal (CC0 1.0) Public Domain
Dedication](https://creativecommons.org/publicdomain/zero/1.0/).
