---
title: "Howard County 2014 Residential Building Permits, Example 2"
author: "Frank Hecker"
date: "February 15, 2015"
output: html_document
---

## Introduction

This is an extension of my previous example of analyzing building
permit data from the `data.howardcountymd.gov` site to answer the
following question: Which localities within Howard County, Maryland,
saw the most residential building permits issued in 2014?

In this version I streamline the analysis by using a data processing
“pipeline” and then graph the results.

## Load libraries

For this and future analyses I’ll again be using the [R statistical
package][] run from the [RStudio][] development environment, along
with the [dplyr package][] to do data manipulation. Since I want to
graph the results I also load the [ggplot2 package][], a plotting
library from the same people who created dplyr.

[R statistical package]: http://r-project.org
[RStudio]: http://www.rstudio.com/products/rstudio
[dplyr package]: http://cran.rstudio.com/web/packages/dplyr/vignettes/introduction.html
[ggplot2 package]: http://ggplot2.org

```{r initialize}
library("dplyr", warn.conflicts = FALSE)
library("ggplot2")
```

## Loading the data

First I download the CVS-format data relating to issuance of building
permits from the `data.howardcountymd.gov` site and store it in a
local file `hoco-building-permits.csv`.

```{r}
download.file("https://data.howardcountymd.gov/geoserver/ows?service=WFS&version=1.0.0&request=GetFeature&typeName=general:Permits_View_Building_New&outputFormat=csv",
              "hoco-building-permits.csv", method = "curl")
```

Next I read the CSV data and convert it into a data frame.

```{r}
permits <- read.csv("hoco-building-permits.csv", stringsAsFactors = FALSE)
```

At the time of writing there are a total of `r nrow(permits)` rows in
the dataset, each representing a single issued building permit.  Since
this dataset gets continually updated as more permits are issued, as
time goes on the number of rows in the dataset will grow.

## The data processing pipeline

Now that I have the permits data I have to do several data processing
steps to select only residential permits for 2014, count the number of
permits per locality and zip code, and sort the data into descending
order by number of residential permits issued.

In the previous example I did this one step at a time, taking one data
set and processing it to produce a new data set, then taking that data
set and processing it to produce yet another data set, and so on until
I had the final data set containg the results. Fortunately the dplyr
package provides a simpler way to specify these steps, using so-called
“pipeline” syntax.

The basic idea is this: Suppose we have a data set `dsinput` with
variables `a`, `b`, `c`, and so on, and we want to transform it into a
second data set `dsoutput` by first selecting only variables `a` and
`b` and then filtering to include only rows where the value of `a` is
equal to 100. In the more traditional approach we would first do the
selection of variables as follows to create a temporary data frame
`dstemp`:

```
dstemp <- select(dsinput, a, b)
```

and then filter `dstemp` to create the final data set `dsoutput`:

```
dsoutput <- filter(dstemp, a == 100)
```

The dplyr pipeline syntax allows us to express the first step as
follows:

```
dstemp <- dsinput %>% select(a, b)
```

The syntax `dsinput %>% select(a, b)` is equivalent to `select(dsinput,
a, b)` but better expresses the idea of a pipeline where the results
of one step are used as input to the next step.

Similarly we can re-write the second step as follows:

```
dsoutput <- dstemp %>% filter(a == 100)
```

where `dstemp %>% filter(a == 100)` is equivalent to `filter(dstemp,
a == 100)`.

But now that we’re using the pipeline syntax like this we no longer
need to explicitly reference the intermediate data frame
`dstemp`. Instead we can express the entire data processing pipeline
in one line:

```
dsoutput <- dsinput %>% select(a, b) %>% filter(a == 100)
```

If we want to add additional data processing steps we can simply
insert them into the pipeline as appropriate.

## The streamlined analysis

In this case I need to do the following processing steps:

* Start with the `permits` data set I just read in.
* Select only the variables of interest.
* Filter the data to include only residential permits.
* Filter the data to include only permits issued in 2014.
* Create a new variable to store the combined locality name and zip
  code.
* Group the permits by locality/zip code.
* Count the number of permits in each group.
* Sort the rows into descending order based on the number of permits.
* Produce a final data set `permits_by_zip`.

Using the pipeline syntax I can compress all of the steps in the above
list into one R statement:

```{r}
permits_by_zip <- permits %>%
    select(Issued_Date, Permit_Type_2, Detailed_Permit_Type, City, Zip) %>%
    filter(Permit_Type_2 == "Residential") %>%
    filter(grepl("/2014$", Issued_Date)) %>%
    mutate(CityZip = paste(City, Zip, sep = "/")) %>%
    group_by(CityZip) %>%
    summarise(Permits = n()) %>%
    arrange(desc(Permits))
```

The new data frame `permits_by_zip` has only `r nrow(permits_by_zip)`
rows, one per locality, and only two fields: `CityZip`, the variable I
grouped by, and `Permits`, the variable containing the number of
permits in each group (returned by the `n()` function).

## Graphing the number of permits by locality

As I did before, I print the entire resulting data frame to show all
of the localities and the number of residential building permits
issued for each one in 2014.

```{r}
print.data.frame(permits_by_zip)
```

However it would also be nice to have a graph of the data, say as a
bar chart, in order to get a better sense of the relative numbers of
permits issued in each locality and zip code. I do that using the
ggplot2 package, which implements a “grammar of graphics” (hence the
name) that is somewhat difficult to learn but once learned makes it
relatively easy to produce professional-looking graphs.

I first start out with a relatively basic graph, a bar chart with the
minimum information specified in order to produce a plot. I use the
`ggplot()` function to specify the data going into the plot (the
`permits_by_zip` data frame) and the “aesthetics” of the plot, that
is, that I want to plot the localities on the x axis and the number of
permits on the y axis:

```{r}
g <- ggplot(permits_by_zip, aes(x = CityZip, y = Permits))
```

The `ggplot()` function returns an object `g` that I can then modify
to produce the actual graph. In this case I want to start with a
simple bar chart, so I modify `g` to include a geometric object or
“geom” consisting of multiple bars (one per locality/zip code) with
the height of the bars equal to the value of the `Permits`
variable. (This is what the `stat = "identity"` expression specifies.)
I then use the `print()` function to actually plot the graph.

```{r}
g <- g + geom_bar(stat = "identity")
print(g)
```

Unfortunately this graph falls short in a number of areas. Most
notably the names of the localities are all run together in a way that
makes them unreadable. I can improve the graph by doing the
following:

* Add more descriptive labels for the x and y axes.
* Add a descriptive title for the graph.
* Most important, reorient the locality names and zip codes so that
  they run vertically rather than horizontally.

I modify the graph object `g` to make these changes and then print it
again. The `xlab()`, `ylab()`, and `ggtitle()` functions should be
self-explanatory. Reorienting the names and zip codes is a bit more
complicated. I confess that I resorted to an online search to find
an [answer on the Stack Overflow developer site][stackoverflow]
that describes how to do this. (Although it’s apparent only in the
underlying source code for this document, I also increased the height
of the overall graph a bit to compensate for the extra space taken up
by the vertical labels.)

[stackoverflow]: http://stackoverflow.com/questions/1330989/rotating-and-spacing-axis-labels-in-ggplot2

```{r fig.height = 6}
g <- g +
    theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)) +
    xlab("Locality/Zip Code") +
    ylab("Number of Permits Issued") +
    ggtitle("Howard County, Maryland, 2014 Residential Building Permits")
print(g)
```

This graph could be further improved in various ways, but as is it
presents the information in a reasonably readable way.

## Conclusion

Using data processing pipelines with the dplyr functions makes it much
easier to understand conceptually what’s going on in the analysis:
each step performs a clearly delineated task, with its output being
the input for the next step. It’s easy to see how the above analysis
could be modified to cover a different year, or to count commercial
building permits rather than residential permits.

The ggplot2 package is a natural complement to dplyr, as both
use data frames as their data structure of choice. In this case
plotting the data makes it much easier to see the wide disparities
among localities in Howard County in terms of residential building
permit issuance.

That concludes this example of analyzing Howard County building permit
data. If I have time to do another example I’ll try to create an
actual map showing the relative number of permits issued per zip code
within the county.

## Appendix

I used the following R environment in doing the analysis for this
example:

```{r sessionInfo}
sessionInfo()
```

You can find the source code for this analysis and others at my
[HoCoData repository on GitHub][hocodata]. This document and its
source code are available for unrestricted use, distribution and
modification under the terms of the Creative Commons [CC0 1.0
Universal (CC0 1.0) Public Domain Dedication][CC0].  Stated more
simply, you’re free to do whatever you‘d like with it.

[hocodata]: https://github.com/frankhecker/hocodata
[CC0]: https://creativecommons.org/publicdomain/zero/1.0/
