---
title: "Creating Howard County Precinct Cartograms Based on 2014 Registered
Voters, Part 1"
author: "Frank Hecker"
date: "March 1, 2015"
output: html_document
---

## Introduction

In doing analyses of precinct-level voting data from the Howard County
2014 general election (using data from the [Maryland State Board of
Elections][MSBoE] and the [Howard County Board of Elections][HCBoE]),
one problem is that maps drawn using that data are inherently
misleading: Precincts in western Howard County are much larger than
precincts in Columbia and eastern Howard County, so they visually
dominate the maps.

[MSBoE]: http://elections.state.md.us/elections/2014/election_data/index.html
[HCBoE]: http://www.howardcountymd.gov/Departments.aspx?id=4294968268

One way to reduce this effect is to create [cartograms][], maps in
which the visual sizes of the geographic subdivisions are based not on
their actual geographic area but on some other variable associated
with them. In particular, for political maps of Howard County I would
like display precincts sized according to the number of registered
voters in each precinct.

[cartograms]: http://en.wikipedia.org/wiki/Cartogram

Creating a voter-based cartogram of Howard County precincts requires
distorting the precinct boundaries so as to change the area of each
precincts while still preserving the overall shape of the precinct as
much as possible, and also preserving its relationship to its
neighboring precincts. This requires some relatively sophisticated
mathematics, and unfortunately there is no existing R package that can
do it well. Instead I use [scapetoad][], a Java-based application
available for Microsoft Windows, Mac OS X, and Linux.

[scapetoad]: http://scapetoad.choros.ch

In this document I describe the overall process of preparing Howard
County map data for use with the scapetoad application. In the next
document (part 2) I describe running scapetoad to produce the desired
cartograms.

## Load packages

I use the [R statistical package][] run from the [RStudio][]
development environment, along with the [dplyr][] package to do data
manipulation and the [sp][] and [rgdal][] packages to read and write
mapping data.

[R statistical package]: http://r-project.org
[RStudio]: http://www.rstudio.com/products/rstudio
[dplyr]: http://cran.rstudio.com/web/packages/dplyr/vignettes/introduction.html
[sp]: http://cran.r-project.org/web/packages/sp/index.html
[rgdal]: http://cran.r-project.org/web/packages/rgdal/index.html

```{r}
library("dplyr", warn.conflicts = FALSE)
library("sp")
library("rgdal")
```

The rgdal package also requires installing the GDAL mapping library on
the underlying operating system.

## General approach

The scapetoad utility works only with map data in ESRI shapefile
format, and requires that the variable used to determine precinct
sizes be included with that data. The scapetoad utility also has a
handy feature whereby you can include other map layers and distort
them in tandem with the original map. I use this to create maps of the
Howard County council districts, Maryland state legislative districts,
and US Congressional districts that match the precinct cartogram.

## Downloading data

First I download a CSV-format dataset containing Howard County 2014
general election turnout statistics by precincts. This dataset
contains the number of registered voters per precinct on election
day. I store a copy of the data in the local file
`hocomd-2014-general-election-turnout-by-precinct.csv`.

```{r}
download.file("https://raw.githubusercontent.com/frankhecker/hocodata/master/datasets/hocomd-2014-general-election-turnout-by-precinct.csv",
              "hocomd-2014-general-election-turnout-by-precinct.csv",
              method = "curl")
```

Then I download spatial data from [`data.howardcountymd.gov`][]
specifying the boundaries of Howard County election precincts, county
council districts, Maryland state legislative districts, and US
Congressional districts. I choose to use the ESRI shapefile format,
and store the data locally in the files `Voting_Precincts.zip`,
`Council_Districts.zip`.  `Congresional_Districts.zip`.

[`data.howardcountymd.gov`]: http://data.howardcountymd.gov

(Notes: The council district map uses the new council boundaries in
effect for the 2014 primary and general elections. The
`data.howardcountymd.gov` site mistakenly lists these boundaries as
not taking effect until December 2014. Also, strictly speaking the
“legislative districts” map shows the House of Delegates districts. A
Maryland State Senate district may be composed of multiple House of
Delegates districts; thus, for example, State Senate district 9 is the
combination of House of Delegates districts 9A and 9B.)

```{r}
download.file("https://data.howardcountymd.gov/geoserver/ows?service=WFS&version=1.0.0&request=GetFeature&typeName=general:Voting_Precincts&outputFormat=shape-zip",
              "Voting_Precincts.zip",
              method = "curl")
download.file("https://data.howardcountymd.gov/geoserver/ows?service=WFS&version=1.0.0&request=GetFeature&typeName=general:Council_Districts&outputFormat=shape-zip",
              "Council_Districts.zip",
              method = "curl")
download.file("https://data.howardcountymd.gov/geoserver/ows?service=WFS&version=1.0.0&request=GetFeature&typeName=general:Legislative_Districts&outputFormat=shape-zip",
              "Legislative_Districts.zip",
              method = "curl")
download.file("https://data.howardcountymd.gov/geoserver/ows?service=WFS&version=1.0.0&request=GetFeature&typeName=general:Congressional_Districts&outputFormat=shape-zip",
              "Congressional_Districts.zip",
              method = "curl")
```

The shapefile data is downloaded in the form of compressed `.zip`
files, so I then unzip the data into the current directory. Each
`.zip` file produces five separate files when uncompressed, with
suffixes `.cst`, `.dbf`, `.prj`, `.shp`, and `.shx`. (They also
produce a sixth file `wfsrequest.txt` that we can ignore.)

```{r}
unzip("Voting_Precincts.zip", overwrite = TRUE)
unzip("Council_Districts.zip", overwrite = TRUE)
unzip("Legislative_Districts.zip", overwrite = TRUE)
unzip("Congressional_Districts.zip", overwrite = TRUE)
```

## Reading the data

First I read in the turnout data.

```{r}
turnout_df <- read.csv("hocomd-2014-general-election-turnout-by-precinct.csv",
                       stringsAsFactors = FALSE)
str(turnout_df)
```

Then I read in the shapefile data for the precinct boundaries.

```{r}
precinct_map <- readOGR(dsn = ".",
                         layer = "Voting_PrecinctsPolygon")
str(precinct_map@data)
```

Now I have to merge the registered voter data in `turnout_df` with the
map data in `precinct_map`. Among other things, this requires
converting the precinct designators to a common format (in this case
‘0-00’, e.g., ‘6-35’).

```{r}
voters_df <- turnout_df %>%
    mutate(PRECINCT20 = gsub("^00([0-9])-0([0-9][0-9])$", "\\1-\\2", Precinct)) %>%
    select(PRECINCT20, Reg.Voters)
str(voters_df)
precinct_map@data <- precinct_map@data %>%
    left_join(voters_df, by = "PRECINCT20")
str(precinct_map@data)
```

Finally I write out a new shapefile for use as input to the scapetoad
utility.

```{r}
writeOGR(precinct_map,
         dsn = ".",
         layer = "Precinct_VotersPolygon",
         driver = "ESRI Shapefile",
         overwrite_layer = TRUE)
```

At this point there is nothing further to be done in R. The next
document (part 2) describes the process of running the scapetoad
application.

## Appendix

I used the following R environment in doing the analysis for this
example:

```{r sessionInfo}
sessionInfo()
```

The underlying GDAL library for the rgdal packages is from the
KyngChaos [GDAL Complete][] distribution version 1.11 for Mac OS X.

[GDAL Complete]: http://www.kyngchaos.com/software/frameworks#gdal_complete

You can find the source code for this analysis and others at my
[HoCoData repository on GitHub][hocodata]. This document and its
source code are available for unrestricted use, distribution and
modification under the terms of the Creative Commons [CC0 1.0
Universal (CC0 1.0) Public Domain Dedication][CC0].  Stated more
simply, you’re free to do whatever you’d like with it.

[hocodata]: https://github.com/frankhecker/hocodata
[CC0]: https://creativecommons.org/publicdomain/zero/1.0/
