---
title: "Creating Howard County Precinct Cartograms Based on 2014
Registered Voters, Part 3"
author: "Frank Hecker"
date: "March 9, 2015"
output: html_document
---

## Introduction

_This is part 3 of a three-part series; see also [part 1][] and
[part 2][]._

[part 1]: http://rpubs.com/frankhecker/63528
[part 2]: http://rpubs.com/frankhecker/63529

In doing analyses of precinct-level voting data from the Howard County
2014 general election (using data from the [Maryland State Board of
Elections][MSBoE] and the [Howard County Board of Elections][HCBoE]),
one problem is that maps drawn using that data are inherently
misleading: Precincts in western Howard County are much larger than
precincts in Columbia and eastern Howard County, so they visually
dominate the maps.

[MSBoE]: http://elections.state.md.us/elections/2014/election_data/index.html
[HCBoE]: http://www.howardcountymd.gov/Departments.aspx?id=4294968268

One way to reduce this effect is to create [cartograms][], maps in
which the visual sizes of the geographic subdivisions are based not on
their actual geographic area but on some other variable associated
with them. In particular, for political maps of Howard County I would
like to display precincts sized according to the number of registered
voters in each precinct.

[cartograms]: http://en.wikipedia.org/wiki/Cartogram

Creating a voter-based cartogram of Howard County precincts requires
distorting the precinct boundaries so as to change the area of each
precinct while still preserving the overall shape of the precinct as
much as possible, and also preserving its relationship to its
neighboring precincts. This requires some relatively sophisticated
mathematics, and unfortunately there is no existing R package that can
do it well. Instead I use [scapetoad][], a Java-based application
available for Microsoft Windows, Mac OS X, and Linux.

[scapetoad]: http://scapetoad.choros.ch

In [part 1][] of this series I described the overall process of
preparing Howard County map data for use with the scapetoad
application. In [part 2][] I described the process of running the
scapetoad application to produce the desired cartogram(s) and checking
the results. In this document I present examples of the completed
cartograms, including overlaying Howard County council district,
Maryland state legislative district, and US congressional district
boundaries on the base precinct-level cartogram. I present an
unmodified precinct map as a comparison.

## Loading libraries

I use the [R statistical package][] run from the [RStudio][]
development environment, along with the [dplyr][] package for data
manipulation, the [ggplot2] package for plotting, and the [sp][] and
[rgdal][] packages to read mapping data.

[R statistical package]: http://r-project.org
[RStudio]: http://www.rstudio.com/products/rstudio
[dplyr]: http://cran.rstudio.com/web/packages/dplyr/vignettes/introduction.html
[ggplot2]: http://ggplot2.org
[sp]: http://cran.r-project.org/web/packages/sp/index.html
[rgdal]: http://cran.r-project.org/web/packages/rgdal/index.html

```{r}
library("dplyr", warn.conflicts = FALSE)
library("ggplot2")
library("sp")
library("rgdal")
```

The rgdal package also requires installing the GDAL mapping library on
the underlying operating system.

## Overall approach

In this document I perform the following steps:

* Download and load all the necessary cartogram shapefiles, along with
  an unmodified precinct map for comparison.
* Transform the cartogram shapefiles and the unmodified precinct map
  for use with the `ggplot()` function.
* Determine where to place labels for individual map areas (precincts
  and districts).
* Plot the unmodified precinct map and the base precinct cartogram
  showing the precincts with labels.
* Plot each of the district cartograms in turn as a layer with labels
  on top of the precinct-level cartogram.

## Downloading the cartogram shapefiles

I first download an unmodified precinct map as published on
[`data.howardcountymd.gov`][].

[`data.howardcountymd.gov`]: http://data.howardcountymd.gov

```{r}
download.file("https://data.howardcountymd.gov/geoserver/ows?service=WFS&version=1.0.0&request=GetFeature&typeName=general:Voting_Precincts&outputFormat=shape-zip",
              "Voting_Precincts.zip",
              method = "curl")
```

I then download all of the cartogram shapefiles as published to the
[datasets][] section of my [hocodata][] GitHub repository.

[datasets]: https://github.com/frankhecker/hocodata/tree/master/datasets
[hocodata]: https://github.com/frankhecker/hocodata

```{r}
url <- "https://raw.githubusercontent.com/frankhecker/hocodata/master/datasets/"
files <- c("Voting_Precincts_Cartogram.zip",
           "Council_Districts_Cartogram.zip",
           "Legislative_Districts_Cartogram.zip",
           "Congressional_Districts_Cartogram.zip")
for (f in files) {
    download.file(paste(url, f, sep = ""), f, method = "curl")
}
```

Since the boundary data is in `.zip` files I need to unzip the files
to extract the actual shapefiles.

```{r}
unzip("Voting_Precincts.zip", overwrite = TRUE)
for (f in files) {
    unzip(f, overwrite = TRUE)
}
```

Finally I read in the boundary shapefile data.

```{r}
unmodified_map <- readOGR(dsn = ".",
                          layer = "Voting_PrecinctsPolygon")
precinct_map <- readOGR(dsn = ".",
                        layer = "Voting_Precincts_CartogramPolygon")
council_map <- readOGR(dsn = ".",
                       layer = "Council_Districts_CartogramPolygon")
legis_map <- readOGR(dsn = ".",
                           layer = "Legislative_Districts_CartogramPolygon")
cong_map <- readOGR(dsn = ".",
                             layer = "Congressional_Districts_CartogramPolygon")
```

## Converting maps for use with `ggplot()`

Before I plot the maps I have to convert the spatial data to
normal data frames usable with the `ggplot()` function.

```{r}
# Define a function to convert spatial data into a ggplot-compatible data frame.
convert_map <- function(map) {
    temp_map <- map  # Avoid modifying original map in the next step
    temp_map@data$id <- rownames(temp_map@data)
    points <- fortify(temp_map, region = "id")
    df <- full_join(points, temp_map@data, by = "id")
    return(df)
}
# Convert the spatial data.
unmodified_df <- convert_map(unmodified_map)
precinct_df <- convert_map(precinct_map)
council_df <- convert_map(council_map)
legis_df <- convert_map(legis_map)
cong_df <- convert_map(cong_map)
```

## Labeling map areas

Since I want to label the various map areas I next compute the
centroids of the precincts and districts in order to position the
labels on the map.

```{r}
# Unmodified precincts
unmodified_centers = coordinates(unmodified_map)
unmodified_centers_df <- as.data.frame(unmodified_centers)
names(unmodified_centers_df) <- c("long", "lat")
unmodified_centers_df$Precinct = as.character(unmodified_map@data$PRECINCT20)
# Precincts
precinct_centers = coordinates(precinct_map)
precinct_centers_df <- as.data.frame(precinct_centers)
names(precinct_centers_df) <- c("long", "lat")
precinct_centers_df$Precinct = as.character(precinct_map@data$PRECINCT20)
# Howard County council districts
council_centers = coordinates(council_map)
council_centers_df <- as.data.frame(council_centers)
names(council_centers_df) <- c("long", "lat")
council_centers_df$District = as.character(council_map@data$DISTRICT20)
# Maryland state legislative districts
legis_centers = coordinates(legis_map)
legis_centers_df <- as.data.frame(legis_centers)
names(legis_centers_df) <- c("long", "lat")
legis_centers_df$District = as.character(legis_map@data$DISTRICT)
# US congressional districts
cong_centers = coordinates(cong_map)
cong_centers_df <- as.data.frame(cong_centers)
names(cong_centers_df) <- c("long", "lat")
cong_centers_df$District = as.character(cong_map@data$Congress_D)
```

The cartogram’s shapes for Howard County council districts 1 and 5 are
so distorted that the centroids are almost outside the district
boundaries. I therefore tweak the locations for the labels for those
districts so that they’ll appear in more suitable locations, moving
the label for district 1 a bit north and the label for district 5 a
bit west.

```{r}
# District 1
council_centers_df$lat[1] <- council_centers_df$lat[1] + 9000
# District 5
council_centers_df$long[5] <- council_centers_df$long[5] - 9000
```

I also tweak the locations for the labels for Maryland state
legislative districts 12, 9A, and 9B so that they’ll appear in more
suitable locations.

```{r}
# District 12
legis_centers_df$long[1] <- legis_centers_df$long[1] + 17400
legis_centers_df$lat[1] <- legis_centers_df$lat[1] - 3000
# District 9A
legis_centers_df$lat[4] <- legis_centers_df$lat[4] + 8400
# District 9B
legis_centers_df$long[3] <- legis_centers_df$long[3] + 9000
```

Finally I tweak the location for the label for US congressional
district 2 so that it will appear in a more suitable location.

```{r}
# District 2
#cong_centers_df$long[1] <- cong_centers_df$long[1] + 17400
cong_centers_df$lat[1] <- cong_centers_df$lat[1] - 6000
```

## Plotting the cartograms

I first plot the unmodified precinct map, as follows:

* Plot the unmodified map for the precincts, outlining their
  boundaries in gray.
* Plot the text labels for the map areas in black at an appropriate
  size.
* Force the x and y coordinate scales to be equal to avoid further
  distortion.
* Remove all extraneous plot elements, leaving only the map
  itself.

I don’t make any effort to tweak the locations of the labels for the
precincts, so labels will appear out of place in cases where the
precincts have odd shapes and the precinct centroids are close to
or even outside the boundaries of the precincts themselves.

```{r fig.width = 9}
g <- ggplot() +
    geom_polygon(data = unmodified_df,
                 aes(x = long, y = lat, group = group),
                 fill = NA,
                 colour = "grey") +
    geom_text(data = unmodified_centers_df,
              aes(x = long, y = lat, label = Precinct),
              size = 2,
              colour = "black",
              show_guide = FALSE) +
    coord_equal() +
    theme(axis.title = element_blank(),
          axis.text = element_blank(),
          axis.ticks = element_blank(),
          panel.grid.major = element_blank(),
          panel.grid.minor = element_blank(),
          panel.background = element_blank())
print(g)
```

I next plot the precinct cartogram, following the same procedure as
for the unmodified precinct map. As with the unmodified map, because
I don’t modify the locations of the text labels some show up in odd
places; the problem is made worse for the cartogram because the
precincts’ shapes are distorted.

```{r fig.width = 9}
g <- ggplot() +
    geom_polygon(data = precinct_df,
                 aes(x = long, y = lat, group = group),
                 fill = NA,
                 colour = "grey") +
    geom_text(data = precinct_centers_df,
              aes(x = long, y = lat, label = Precinct),
              size = 2,
              colour = "black",
              show_guide = FALSE) +
    coord_equal() +
    theme(axis.title = element_blank(),
          axis.text = element_blank(),
          axis.ticks = element_blank(),
          panel.grid.major = element_blank(),
          panel.grid.minor = element_blank(),
          panel.background = element_blank())
print(g)
```

I next plot the Howard County council districts overlaid on the
precinct cartogram. The procedure is the same as in the previous
plots, except that I add a second layer to the plot showing the
council boundaries and I label the council districts instead of the
precincts.

```{r fig.width = 9}
g <- ggplot() +
    geom_polygon(data = precinct_df,
                 aes(x = long, y = lat, group = group),
                 fill = NA,
                 colour = "grey") +
    geom_polygon(data = council_df,
                 aes(x = long, y = lat, group = group),
                 fill = NA,
                 colour = "black") +
    geom_text(data = council_centers_df,
              aes(x = long, y = lat, label = District),
              size = 7,
              colour = "black",
              show_guide = FALSE) +
    coord_equal() +
    theme(axis.title = element_blank(),
          axis.text = element_blank(),
          axis.ticks = element_blank(),
          panel.grid.major = element_blank(),
          panel.grid.minor = element_blank(),
          panel.background = element_blank())
print(g)
```

I plot the Maryland state legislative districts overlaid on the
precinct cartogram.

```{r fig.width = 9}
g <- ggplot() +
    geom_polygon(data = precinct_df,
                 aes(x = long, y = lat, group = group),
                 fill = NA,
                 colour = "grey") +
    geom_polygon(data = legis_df,
                 aes(x = long, y = lat, group = group),
                 fill = NA,
                 colour = "black") +
    geom_text(data = legis_centers_df,
              aes(x = long, y = lat, label = District),
              size = 7,
              colour = "black",
              show_guide = FALSE) +
    coord_equal() +
    theme(axis.title = element_blank(),
          axis.text = element_blank(),
          axis.ticks = element_blank(),
          panel.grid.major = element_blank(),
          panel.grid.minor = element_blank(),
          panel.background = element_blank())
print(g)
```

Finally, I plot the US congressional districts overlaid on the
precinct cartogram.

```{r fig.width = 9}
g <- ggplot() +
    geom_polygon(data = precinct_df,
                 aes(x = long, y = lat, group = group),
                 fill = NA,
                 colour = "grey") +
    geom_polygon(data = cong_df,
                 aes(x = long, y = lat, group = group),
                 fill = NA,
                 colour = "black") +
    geom_text(data = cong_centers_df,
              aes(x = long, y = lat, label = District),
              size = 7,
              colour = "black",
              show_guide = FALSE) +
    coord_equal() +
    theme(axis.title = element_blank(),
          axis.text = element_blank(),
          axis.ticks = element_blank(),
          panel.grid.major = element_blank(),
          panel.grid.minor = element_blank(),
          panel.background = element_blank())
print(g)
```

## Appendix

I used the following R environment in creating this document:

```{r sessionInfo}
sessionInfo()
```

The underlying GDAL library for the rgdal packages is from the
KyngChaos [GDAL Complete][] distribution version 1.11 for Mac OS X.

[GDAL Complete]: http://www.kyngchaos.com/software/frameworks#gdal_complete

You can find the source code for this document and others at my
[HoCoData repository on GitHub][hocodata]. This document and its
source code are available for unrestricted use, distribution and
modification under the terms of the Creative Commons [CC0 1.0
Universal (CC0 1.0) Public Domain Dedication][CC0].  Stated more
simply, you’re free to do whatever you’d like with it.

[hocodata]: https://github.com/frankhecker/hocodata
[CC0]: https://creativecommons.org/publicdomain/zero/1.0/
