---
title: "Howard County Turnout in the 2014 General Election, Part 1"
author: "Frank Hecker"
date: "April 26, 2015"
output: html_document
---

## Introduction

In the 2014 gubermatorial general election in Howard County, Maryland,
a subject of interest was relative turnout for Republican
vs. Democratic voters, with higher Republican turnout seen as the key
for Allan Kittleman’s election as County Executive (not to mention
other Republican victories from Larry Hogan on down). A related
question is whether turnout was depressed in certain Howard County
council districts due to lack of opposition to the incumbent council
members. (In particular, in districts 3 and 4 Jen Terrasa and Mary-Kay
Sigaty respectively had no declared Republican opponents.)

In this and subsequent parts of this series I take a look at turnout
statistics to explore these questions, starting in part 1 with a look
at the state-wide turnout datasets for the Maryland 2014 general
election.

## Load packages

For this analysis I use the [R statistical package][] run from the
[RStudio][] development environment, along with the [readxl][] package
to extract data from Excel spreadsheets and the [dplyr][] package to
do data manipulation.

[R statistical package]: http://r-project.org
[RStudio]: http://www.rstudio.com/products/rstudio
[readxl]: http://cran.rstudio.com/web/packages/readxl/
[dplyr]: http://cran.rstudio.com/web/packages/dplyr/vignettes/introduction.html

```{r}
library("readxl")
library("dplyr", warn.conflicts = FALSE)
```

## Data sources

The Maryland State Board of Elections has published a number of
turnout-related reports (in both PDF and Microsoft Excel format) as
part of its [2014 general election reports][md-ge2014]. The two
reports of most interest for my purpose are the [state-wide per-county
turnout statistics][md-turnout-county] [Excel] and the [state-wide
per-precinct turnout statistics][md-turnout-precinct] [Excel].

[md-turnout-county]: http://www.elections.state.md.us/elections/2014/turnout/general/GG14_Turnout_by_party_by_county.xlsx
[md-turnout-precinct]: http://www.elections.state.md.us/elections/2014/turnout/general/GG14_Turnout_by_party_by_precinct.xlsx
[md-ge2014]: http://www.elections.state.md.us/elections/2014/

First I download the Excel-format data files from the Maryland State
Board of Elections containing statewide 2014 general election turnout
by county and by precincts, and store a copy of the data in the local
files `GG14_Turnout_by_party_by_county.xlsx` and
`GG14_Turnout_by_party_by_precinct.xlsx`.

```{r}
download.file(
"http://www.elections.state.md.us/elections/2014/turnout/general/GG14_Turnout_by_party_by_county.xlsx",
              "GG14_Turnout_by_party_by_county.xlsx",
              method = "curl")
download.file(
"http://www.elections.state.md.us/elections/2014/turnout/general/GG14_Turnout_by_party_by_precinct.xlsx",
              "GG14_Turnout_by_party_by_precinct.xlsx",
              method = "curl")
```

I then load the Excel spreadsheets into the dataframes
`county_turnout_raw` and `precinct_turnout_raw` respectively.

```{r}
county_turnout_raw <- read_excel("GG14_Turnout_by_party_by_county.xlsx")
colnames(county_turnout_raw)[7] <- "Percent_Turnout"
str(county_turnout_raw)
precinct_turnout_raw <- read_excel("GG14_Turnout_by_party_by_precinct.xlsx")
str(precinct_turnout_raw)
```

## Cleaning the data

Unfortunately the spreadsheet for the per-county turnout organizes its
data in a way that is not optimal for analysis in R: There are
separate sections (groups of rows) for each party, each with a header
row and a subtotal row, and an initial section for overall turnout
(all parties).

For my purposes it is better to identify the party using a separate
variable (column), and to eliminate all subtotal and total rows.
(They can always be recomputed as needed.) I therefore do the
following:

* Add a new "Party" column to the dataframe, using as its values a
  vector that has party labels for those rows representing actual
  party turnout (as determined by inspecting the spreadsheet by hand),
  and NA values for other rows (e.g., subtotals, headers, and blank
  rows).
* Retain only rows with a valid party label.

I also convert the various counts from character strings to integers.

```{r}
parties <- c(rep(NA, 26),
             rep("Republican", 24),
             rep(NA, 4),
             rep("Democrat", 24),
             rep(NA, 4),
             rep("Libertarian", 24),
             rep(NA, 4),
             rep("Green", 24),
             rep(NA, 2),
             rep("Other Parties", 24),
             rep(NA, 3),
             rep("Unaffiliated", 24),
             rep(NA, 6))
county_turnout <- county_turnout_raw %>%
    mutate(Party = parties) %>%
    filter(!is.na(Party)) %>%
    mutate(LBE = Totals,
           Polls = as.integer(POLLS),
           Early_Voting = as.integer(EV),
           Absentee = as.integer(ABS),
           Provisional = as.integer(PROV),
           Eligible_Voters = as.integer(ELIGIBLE_VOTERS)) %>%
    select(LBE,
           Polls,
           Early_Voting,
           Absentee,
           Provisional,
           Eligible_Voters,
           Percent_Turnout,
           Party)
str(county_turnout)
```

Next I do a check to confirm that the `county_turnout` dataset has
values consistent with the original spreadsheet, by totaling the
counts and confirming that they match the corresponding state-wide
totals from row 25 of the original spreadsheet.

```{r}
count_total_county <- county_turnout %>%
    summarise(Polls = sum(Polls),
              Early_Voting = sum(Early_Voting),
              Absentee = sum(Absentee),
              Provisional = sum(Provisional),
              Eligible_Voters = sum(Eligible_Voters))
print(count_total_county)
stopifnot(count_total_county == county_turnout_raw[25,2:6])
```

Next I turn to the precinct-level turnout data. This dataset is much
easier to work with: Each row contains data for a particular party in
a particular precinct. I change the variable names to mixed case for
readability, the variable types as appropriate, and the values for the
`Party` variable to use initial capitals instead of upper case only.

```{r}
precinct_turnout <- precinct_turnout_raw %>%
    mutate(Election_Code = ELECTION_CODE,
           LBE_Code = as.integer(LBE_CODE),
           Congressional_District_Code = as.integer(CONGRESSIONAL_DISTRICT_CODE),
           Legislative_District_Code = LEGISLATIVE_DISTRICT_CODE,
           Precinct = PRECINCT,
           Party = gsub("(^|[[:space:]])([[:alpha:]])([[:alpha:]]*)",
                        "\\1\\U\\2\\L\\3",
                        PARTY,
                        perl=TRUE),
           Polls = as.integer(POLLS),
           Early_Voting = as.integer(EARLY_VOTING),
           Absentee = as.integer(ABSENTEE),
           Provisional = as.integer(PROVISIONAL),
           Eligible_Voters = as.integer(ELIGIBLE_VOTERS)) %>%
    select(Election_Code,
           LBE,
           LBE_Code,
           Congressional_District_Code,
           Legislative_District_Code,
           Precinct,
           Party,
           Polls,
           Early_Voting,
           Absentee,
           Provisional,
           Eligible_Voters,
           Percentage)
str(precinct_turnout)
```

# Checking the data

In theory the county-level dataset is simply a summary of the
precinct-level dataset. Now that I’ve cleaned the two datasets and
made them consistent in terms of variable names, types, and values I
check that hypothesis.

I first check that the parties listed in the county turnout dataset
are the exact same parties as in the precinct turnout dataset, and
vice versa:

```{r}
sort(unique(county_turnout$Party))
sort(unique(precinct_turnout$Party))
```

Next I check to see if the counts in the precinct turnout dataset are
consistent with the counts in the county turnout dataset, by comparing
the total counts for the entire state calculated using each dataset:

```{r}
count_total_precinct <- precinct_turnout %>%
    summarise(Polls = sum(Polls),
              Early_Voting = sum(Early_Voting),
              Absentee = sum(Absentee),
              Provisional = sum(Provisional),
              Eligible_Voters = sum(Eligible_Voters))
print(count_total_county)
print(count_total_precinct)
```

All counts match except for the total provisional votes, for which
the precinct-level data shows one less provisional vote than the
county-level data.

To narrow down the source of the discrepancy I compare the counts of
provisional votes for each county using the two datasets:

```{r}
provisional_county <- county_turnout %>%
    group_by(LBE) %>%
    summarise(Provisional_County = sum(Provisional))
provisional_precinct <- precinct_turnout %>%
    group_by(LBE) %>%
    summarise(Provisional_Precinct = sum(Provisional))
left_join(provisional_county, provisional_precinct) %>%
    print.data.frame()
```

It appears that the one-vote discrepancy is for Prince George’s
County. To narrow things down even further I print the number of
provisional voters for each party in Prince George’s County:

```{r}
pg_provisional_county <- county_turnout %>%
    filter(LBE == "Prince George\'s") %>%
    mutate(Provisional_County = Provisional) %>%
    select(Party, Provisional_County)
pg_provisional_precinct <- precinct_turnout %>%
    filter(LBE == "Prince George\'s") %>%
    group_by(Party) %>%
    summarise(Provisional_Precinct = sum(Provisional))
left_join(pg_provisional_county, pg_provisional_precinct) %>%
    print.data.frame()
```

The discrepancy is thus due to a difference in the number of
provisional votes cast by members of the Libertarian party in Prince
George’s County.

Next I check the calculation of the turnout percentage figures (which
are actually fractions, not percentages per se). I recalculate the
percentages for the precinct-level data, compute the absolute
difference for each precinct between the original figure and the
recalculated figure, and determine the maximum absolute difference
across all precincts.

(The calculation of turnout percentage is not totally straightforward,
since I must account for cases in which there are no eligible voters
for a given party in a given precinct. In the original dataset the
turnout percentage is set to zero in these cases, so I do the same.)

```{r}
precinct_turnout %>%
    mutate(Pct = ifelse(Eligible_Voters <= 0,
                        0,
                        (Polls+Early_Voting+Absentee+Provisional)/Eligible_Voters)) %>%
    summarise(max(abs(Percentage - Pct)))
```

The maximum absolute difference is zero, so the original percentage
values were indeed calculated as I expected.

## The final precinct-level dataset

With the exception of the one-vote discrepancy in Prince George’s
County noted above, the county-level dataset is simply a summarization
of the precinct-level dataset. I therefore choose to keep only the
precinct-level dataset, and write it out as a CSV file
`gg14-turnout-by-party-by-precinct.csv` for use in the subsequent
parts of this analysis. (I use a slightly different name than that
of the original Excel spreadsheet to emphasize that this file is
not simply a CSV export of the spreadsheet.)

```{r}
write.csv(precinct_turnout,
          file = "gg14-turnout-by-party-by-precinct.csv",
          quote = TRUE,
          row.names = FALSE)
```

The precinct-level dataset contains one row for each combination of
county, precinct, and party; each row contains the following
variables:

* `Election_Code`: The code for the year and type of election; for
  this dataset it has the single value “GG14” (gubernatorial general
  election for 2014). (character string)
* `LBE`: The jurisdiction of the local board of elections (e.g., “Howard”
  or “Baltimore City”). (character string)
* `LBE_Code`: The code for the local board of elections (e.g., 14 for
  Howard County). (integer)
* `Congressional_District_Code`: The US Congressional district with
  which the precinct is associated (e.g., 7). (integer)
* `Legislative_District_Code`: The Maryland state legislative district
  with which the precinct is associated (e.g., “09A” for district 9A).
  It consists of a two-digit zero-filled value optionally followed by
  a single letter “A”, “B”, etc. (character string)
* `Precinct`: The designator for the precinct (e.g., “002-006”). It
  consists of two three-digit zero-filled values separated by a
  hyphen. A precinct designator is unique only within a given
  LBE jurisdiction. (character string)
* `Party`: The party affiliation of people voting, or “Unaffiliated”
  for those who did not indicate a party affiliation when registering
  to vote. The recognized party affiliations are “Republican”,
  “Democrat”, “Libertarian”, and “Green”; several small parties are
  lumped together as “Other Parties”. (character string)
* `Polls`: The number of people who voted at the polls on election
  day. (integer)
* `Early_Voting`: The number of people who voted during the early
  voting period. (integer)
* `Absentee`: The number of people who voted via absentee
  ballot. (integer)
* `Provisional`: The number of people who cast provisional
  ballots. (integer)
* `Eligible_Voters`: The number of people eligible to vote (i.e.,
  registered to vote and otherwise qualified). (integer)
* `Percentage`: The fraction of eligible voters who actually
  voted (e.g., 0.5745). It is computed as the sum of the `Poll`,
  `Early_Voting`, `Absentee`, and `Provisional` counts, divided by the
  `Eligible_Voters` count. Note that it must be multiplied by 100 to
  obtain a true percentage. (numeric)

To make the file `gg14-turnout-by-party-by-precinct.csv` more
accessible to myself and others, I put a copy of the file in my
[hocodata Github repository][]; it can be downloaded at the following
URL:

https://raw.githubusercontent.com/frankhecker/hocodata/master/datasets/gg14-turnout-by-party-by-precinct.csv

[hocodata Github repository]: https://github.com/frankhecker/hocodata/

## Appendix

I used the following R environment in doing the analysis for this
example:

```{r sessionInfo}
sessionInfo()
```

You can find the source code for this analysis and others at my
[HoCoData repository on GitHub][hocodata]. This document and its
source code are available for unrestricted use, distribution and
modification under the terms of the Creative Commons [CC0 1.0
Universal (CC0 1.0) Public Domain Dedication][CC0].  Stated more
simply, you’re free to do whatever you’d like with it.

[hocodata]: https://github.com/frankhecker/hocodata
[CC0]: https://creativecommons.org/publicdomain/zero/1.0/
