---
title: "Creating Howard County Precinct Cartograms Based on 2014 Registered
Voters, Part 2"
author: "Frank Hecker"
date: "March 1, 2015"
output: html_document
---

## Introduction

In doing analyses of precinct-level voting data from the Howard County
2014 general election (using data from the [Maryland State Board of
Elections][MSBoE] and the [Howard County Board of Elections][HCBoE]),
one problem is that maps drawn using that data are inherently
misleading: Precincts in western Howard County are much larger than
precincts in Columbia and eastern Howard County, so they visually
dominate the maps.

[MSBoE]: http://elections.state.md.us/elections/2014/election_data/index.html
[HCBoE]: http://www.howardcountymd.gov/Departments.aspx?id=4294968268

One way to reduce this effect is to create [cartograms][], maps in
which the visual sizes of the geographic subdivisions are based not on
their actual geographic area but on some other variable associated
with them. In particular, for political maps of Howard County I would
like to display precincts sized according to the number of registered
voters in each precinct.

[cartograms]: http://en.wikipedia.org/wiki/Cartogram

Creating a voter-based cartogram of Howard County precincts requires
distorting the precinct boundaries so as to change the area of each
precinct while still preserving the overall shape of the precinct as
much as possible, and also preserving its relationship to its
neighboring precincts. This requires some relatively sophisticated
mathematics, and unfortunately there is no existing R package that can
do it well. Instead I use [scapetoad][], a Java-based application
available for Microsoft Windows, Mac OS X, and Linux.

[scapetoad]: http://scapetoad.choros.ch

In a previous document (part 1) I described the overall process of
preparing Howard County map data for use with the scapetoad
application. In this document I describe the process of running the
scapetoad application to produce the desired cartogram(s) and checking
the results.

## Loading libraries

I use the [R statistical package][] run from the [RStudio][]
development environment, along with the [sp][] and [rgdal][] packages
to read mapping data.

[R statistical package]: http://r-project.org
[RStudio]: http://www.rstudio.com/products/rstudio
[sp]: http://cran.r-project.org/web/packages/sp/index.html
[rgdal]: http://cran.r-project.org/web/packages/rgdal/index.html

```{r}
library("sp")
library("rgdal")
```

The rgdal package also requires installing the GDAL mapping library on
the underlying operating system.

## General approach

The scapetoad application works only with map data in ESRI shapefile
format, and requires that the variable used to determine precinct
sizes be included with that data. The scapetoad application also has a
handy feature whereby you can include other map layers and distort
them in tandem with the original map. I use this to create maps of the
Howard County council districts, Maryland state legislative districts,
and US Congressional districts that match the precinct cartogram.

## Running the scapetoad utility

In the previous document (part 1) I downloaded the shapefile for the
boundaries of Howard County precincts and created a new shapefile
`Precinct_VotersPolygon.shp` that has an added field `Reg.Voters`
specifying the number of registered voters in each precinct as of the
2014 general election. This is the shapefile I use to create the main
cartogram.

I also downloaded shapefiles for the boundaries of Howard County
council districts and for the Howard County portions of Maryland state
legislative districts and US congressional districts. I use these
shapefiles to create additional cartograms.

I then invoke the scapetoad application and use the “Add a Layer”
function multiple times to add invidual layers for the following
shapefiles:

* `Precinct_VotersPolygon.shp`
* `Council_DistrictsPolygon.shp`
* `Legislative_DistrictsPolygon.shp`
* `Congressional_DistrictsPolygon.shp`

I next use the ‘Create Cartogram’ function to create the various
cartograms. I do the following steps in the resulting wizard:

* Click the ‘Next’ button to get past the first screen.
* Select ‘Precinct_VotersPolygon’ for spatial coverage (the main layer
  to be transformed), then click the ‘Next’ button.
* Select ‘Reg_Voters’ for the cartogram attribute (the metric unit of
  the new space). (Note that the scapetoad application changes the
  field name from the original ‘Reg.Voters’.) Then click the ‘Next’
  button.
* Click the ‘Layers to transform...’ button and select all three
  layers presented: ‘Congressional_DistrictsPolygon’,
  ‘Legislative_DistrictsPolygon’, and ‘Council_DistrictsPolygon’.
  These layers will be transformed based on the transformations
  applied to the main layer (‘Precinct_VotersPolygon’). Then click
  the ‘Next’ button.
* Move the slider to select ‘High’ quality, then click the ‘Compute’
  button. Wait for the transformation process to complete; it takes
  several minutes. Then click the ‘Done’ button.

Finally I use the ‘Export to shape’ function multiple times to export
each of the transformed layers to a shapefile:

* Select layer ‘Precinct_VotersPolygon (2)’, the transformed layer for
  the precinct boundaries. (‘Precinct_VotersPolygon’ is the original
  untransformed layer.) Click on the ‘OK’ button and save the layer as
  ‘Voting_Precincts_CartogramPolygon’.
* Select layer ‘Council_DistrictsPolygon (2)’ and save it as
  ‘Council_Districts_CartogramPolygon’.
* Select layer ‘Legislative_DistrictsPolygon (2)’ and save it as
  ‘Legislative_Districts_CartogramPolygon’.
* Select layer ‘Congressional_DistrictsPolygon (2)’ and save it as
  ‘Congressional_Districts_CartogramPolygon’.

Each export operation produces three files, with suffixes `.dbx`,
`.shp`, and `.shx`.

At this point I am finished running the scapetoad application.

## Checking the resulting cartograms

I now check the newly-created cartograms against the original maps. I
read in the shapefile data for the cartograms for the precincts,
county council districts, Maryland state legislative districts, and US
Congressional districts, along with the shapefiles for the original
maps as downloaded from `data.howardcountymd.gov`.

```{r}
precinct_cg <- readOGR(dsn = ".",
                       layer = "Voting_Precincts_CartogramPolygon")
precinct_map <- readOGR(dsn = ".",
                        layer = "Voting_PrecinctsPolygon")
council_cg <- readOGR(dsn = ".",
                      layer = "Council_Districts_CartogramPolygon")
council_map <- readOGR(dsn = ".",
                       layer = "Council_DistrictsPolygon")
leg_cg <- readOGR(dsn = ".",
                  layer = "Legislative_Districts_CartogramPolygon")
leg_map <- readOGR(dsn = ".",
                   layer = "Legislative_DistrictsPolygon")
cong_cg <- readOGR(dsn = ".",
                   layer = "Congressional_Districts_CartogramPolygon")
cong_map <- readOGR(dsn = ".",
                    layer = "Congressional_DistrictsPolygon")
```

Then I check the fields in the precinct cartogram vs. the fields in
the original map:

```{r}
str(precinct_map@data)
str(precinct_cg@data)
```

Note the added `Reg_Voters` field from the turnout statistics. (The
other two fields `Reg_Voters_D` and `SizeError` are created by the
scapetoad application.)

Now I do quick plots of the original maps and the cartograms. I use
the base `plot()` function in R to avoid the need to create dataframes
for the `ggplot()` function.

```{r}
plot(precinct_map)
plot(precinct_cg)
plot(council_map)
plot(council_cg)
plot(leg_map)
plot(leg_cg)
plot(cong_map)
plot(cong_cg)
```

As a final step I create new `.zip` files from the shapefiles created
by the scapetoad application.

```{r}
files <- c("Voting_Precincts",
           "Council_Districts",
           "Legislative_Districts",
           "Congressional_Districts")
suffixes <- c(".dbf", ".shp", ".shx")
for (file in files) {
        zip_file <- paste(file, "_Cartogram.zip", sep = "")
        map_files <- paste(file, "_CartogramPolygon", suffixes, sep = "")
        if (file.exists(zip_file)) file.remove(zip_file)
        zip(zip_file, map_files)
}
```

I can now use these `.zip` files in place of the original shapefile
`.zip` files downloaded from `data.howardcountymd.gov`.

## Appendix

I used the following R environment in doing the analysis for this
example:

```{r sessionInfo}
sessionInfo()
```

The underlying GDAL library for the rgdal packages is from the
KyngChaos [GDAL Complete][] distribution version 1.11 for Mac OS X.

[GDAL Complete]: http://www.kyngchaos.com/software/frameworks#gdal_complete

You can find the source code for this analysis and others at my
[HoCoData repository on GitHub][hocodata]. This document and its
source code are available for unrestricted use, distribution and
modification under the terms of the Creative Commons [CC0 1.0
Universal (CC0 1.0) Public Domain Dedication][CC0].  Stated more
simply, you’re free to do whatever you’d like with it.

[hocodata]: https://github.com/frankhecker/hocodata
[CC0]: https://creativecommons.org/publicdomain/zero/1.0/
