---
title: "Allan Kittleman’s Election-Day Victory Margins in the Howard
County 2014 General Election, Part 2"
author: "Frank Hecker"
date: "February 28, 2015"
output: html_document
---

## Introduction

In this analysis I use precinct-level results from the Howard County
2014 general election (courtesy of the [Maryland State Board of
Elections][MSBoE]) to look at Allan Kittleman’s margin of victory
across the county on election day in the race for Howard County
Executive.  I’m interested in the general question of whether there
was an “enthusiasm gap” in which Kittleman’s election-day results were
particularly lopsided, e.g., due to increased turnout of Republican
voters or unusually high support for Kittleman from Democrats and
unaffiliated voters.

[MSBoE]: http://elections.state.md.us/elections/2014/election_data/index.html

I present the data as a map of Howard County with the precincts
colored according to Kittleman’s absolute and relative margins of
victory, and with county council boundaries added. The map is based on
precinct and council boundaries made available by the Howard County
GIS division on the [data.howardcountymd.gov][] site.

[data.howardcountymd.gov]: http://data.howardcountymd.gov/

## Load packages

For this analysis I use the [R statistical package][] run from the
[RStudio][] development environment, along with the [dplyr][] and
[tidyr][] packages to do data manipulation and the [ggplot2][] package
to draw the histogram and map.

[R statistical package]: http://r-project.org
[RStudio]: http://www.rstudio.com/products/rstudio
[dplyr]: http://cran.rstudio.com/web/packages/dplyr/vignettes/introduction.html
[tidyr]: http://blog.rstudio.org/2014/07/22/introducing-tidyr/
[ggplot2]: http://ggplot2.org

```{r}
library("dplyr", warn.conflicts = FALSE)
library("tidyr")
library("ggplot2")
```

I also need to load R packages used to manipulate spatial data in R. I
first load the [sp][] package, a prerequisite for using other spatial
data packages. I use the [rgdal][] package to load spatial data for
boundaries downloaded from `data.howardcountymd.gov`.

[sp]: http://cran.r-project.org/web/packages/sp/index.html
[rgdal]: http://cran.r-project.org/web/packages/rgdal/index.html

```{r}
library("sp")
library("rgdal")
```

The rgdal package also requires installing the GDAL mapping library on
the underlying operating system.

## General approach

How would one best measure relative voter enthusiasm for Allan
Kittleman vs. Courtney Watson? One measure would be how each candidate
outperformed their “expected” vote, for example, how many votes
Kittleman attracted in a given precinct vs. the number of registered
Republicans in that precinct, and ditto for Watson vis-a-vis the
number of registered Democrats. A related measure would look at
Republican turnout (i.e., as a percentage of registered Republicans)
in a given precinct vs. Democratic turnout.

In this document I confine myself to looking at simple margins of
victory in each precinct. The Maryland State Board of Elections has
now made available precinct-level data (in Microsoft Excel format)
giving party turnout in the 2014 general election. I’ll take a look at
that data in a later analysis.

As I mentioned above, this analysis is for election day voting
only. Absentee ballots and votes cast at early voting centers are
_not_ included in the per-precinct totals as reported by the Maryland
State Board of Elections. I’m not aware of any good method to assign
absentee and early voting results to individual precincts.

## Loading the data

First I download the CVS-format data file from the Maryland State
Board of Elections containing Howard County 2014 general election
results by precincts, and store a copy of the data in the local file
`Howard_By_Precinct_2014_General.csv`.

```{r}
download.file("http://elections.state.md.us/elections/2014/election_data/Howard_By_Precinct_2014_General.csv",
              "Howard_By_Precinct_2014_General.csv",
              method = "curl")
```

Then I download spatial data from `data.howardcountymd.gov` specifying
the boundaries of Howard County election precincts and county council
districts, using the new council boundaries in effect for the 2014
primary and general elections). (The `data.howardcountymd.gov` site
mistakenly lists these boundaries as not taking effect until December
2014.) I choose to use the GeoJSON format, and store the data locally
in the files `precincts.json` and `districts.json`.

```{r}
download.file("https://data.howardcountymd.gov/geoserver/ows?service=WFS&version=1.0.0&request=GetFeature&typeName=general:Voting_Precincts&outputFormat=application/json",
              "precincts.json",
              method = "curl")
download.file("https://data.howardcountymd.gov/geoserver/ows?service=WFS&version=1.0.0&request=GetFeature&typeName=general:Council_Districts&outputFormat=application/json",
              "districts.json",
              method = "curl")
```

I then read in the CSV file for election results. I remove extraneous
spaces from the names of the offices to make it easier to filter the
results by office.

```{r}
hoco_ge14_df <- read.csv("Howard_By_Precinct_2014_General.csv", stringsAsFactors = FALSE)
hoco_ge14_df$Office.Name <- gsub("  *$", "", hoco_ge14_df$Office.Name)
```

Finally I read in the GeoJSON precinct and council district boundary
data.

```{r}
precincts_spdf <- readOGR("precincts.json", "OGRGeoJSON")
districts_spdf <- readOGR("districts.json", "OGRGeoJSON")
```

Recall that the `readOGR()` function returns a special type of data
structure, a “SpatialPolygonsDataFrame” containing both a regular data
frame and a list of polygons containing map data for the precincts.

## Data processing

The processing of the data is identical to that done for the last
example I did analyzing Allan Kittleman’s victory margins in the
County Executive race.  For brevity I consolidate everything into a
single data processing pipeline. The operations in the pipeline are as
follows:

* Filter the rows based on `Office.Name` to get the results of the
  County Executive race.
* Select only the columns of interest.
* Filter out write-in votes.
* Combine the `Election.District` and `Election.Precinct` variables
  into a single variable `Precinct` having the form ‘0-00’.
* Take the values of the `Party` variable in different rows and
  convert them into column variables `REP` and `DEM`, taking the
  values from the `Election.Night.Votes` variable.
* Compute a `REP.Margin` variable containing the absolute Republican
  margin of victory and a `Pct.REP.Margin` variable containing the
  Republican margin of victory in percentage terms (rounded to one
  digit past the decimal place).

```{r}
ak_margins_df <- hoco_ge14_df %>%
    filter(Office.Name == "County Executive") %>%
    select(Election.District, Election.Precinct, Party, Election.Night.Votes) %>%
    filter(Party != "BOT") %>%
    mutate(Precinct = paste(as.character(Election.District),
                            "-",
                            formatC(Election.Precinct, width = 2, flag = 0),
                            sep = "")) %>%
    select(-Election.District, -Election.Precinct) %>%
    spread(Party, Election.Night.Votes) %>%
    mutate(REP.Margin = REP - DEM,
           Pct.REP.Margin = round(100 * (REP - DEM) / (REP + DEM), 1))
```

As a check I print the five precincts in which Allan Kittleman
received his highest margins on election day in terms of absolute
votes:

```{r}
ak_margins_df %>% arrange(desc(REP.Margin)) %>% head(5)
```

and the five precincts in which Courtney Watson received her highest
margins on election day in terms of absolute votes:

```{r}
ak_margins_df %>% arrange(REP.Margin) %>% head(5)
```

I also print summary statistics for the entire data set:

```{r}
summary(ak_margins_df)
```

Among other things this gives ranges for Kittleman’s margins in terms
of votes (`r min(ak_margins_df$REP.Margin)` to
`r max(ak_margins_df$REP.Margin)`) and percentages
(`r min(ak_margins_df$Pct.REP.Margin)`% to
`r max(ak_margins_df$Pct.REP.Margin)`%). I use these later when
assigning colors to the precincts based on Kittleman’s absolute or
percentage margins of victory.

## Creating precinct-level maps

Now comes the fun part: actually mapping the data.  First I convert
the precincts and council district map data to normal data frames
usable with the `ggplot()` function.

```{r}
precincts_spdf@data$id <- rownames(precincts_spdf@data)
precincts_points <- fortify(precincts_spdf, region = "id")
precincts_df <- full_join(precincts_points, precincts_spdf@data, by = "id")
districts_spdf@data$id <- rownames(districts_spdf@data)
districts_points <- fortify(districts_spdf, region = "id")
districts_df <- full_join(districts_points, districts_spdf@data, by = "id")
```

Then I add the margins data to the precinct map data.

```{r}
precincts_df <- precincts_df %>%
    mutate(Precinct = as.character(PRECINCT20)) %>%
    left_join(ak_margins_df, by = "Precinct")
```

Since I want to label the county council districts I next compute the
centroids of the districts in order to position the labels on the map.

```{r}
district_centers = coordinates(districts_spdf)
district_centers_df <- as.data.frame(district_centers)
names(district_centers_df) <- c("long", "lat")
district_centers_df$District = as.character(districts_spdf@data$DISTRICT20)
```

Next I plot Allan Kittleman’s victory margins by precinct, starting
with the absolute margins in votes. This plot contains three layers:

* Polygons for the precincts, with a fill color set for each polygon
  according to Kittleman’s victory margin in that precinct, using the
  conventional color scale of red for Republican-leaning precincts and
  blue for Democratic-leaning precincts.
* Polygons for the county council districts, outlined in white with no
  fill color set (i.e., to make the polygons transparent).
* Text labels for the county council districts.

I also tweak the plot as follows:

* Kittleman’s maximum victory margin
  (`r max(ak_margins_df$REP.Margin)` votes) is significantly higher
  than Courtney Watson’s maximum victory margin
  (`r abs(min(ak_margins_df$REP.Margin))` votes). I therefore set the
  coloring to have maximum red at a 700 vote margin and maximum blue
  at -700. This treats the two candidates’ margins equally in terms of
  the coloring.
* I set the x and y coordinate scales to be equal to avoid having the
  map be distorted horizontally or vertically.
* I remove all text on the x and y axes.
* I add a title for the graph.

```{r fig.width = 9}
g <- ggplot() +
    geom_polygon(data = precincts_df,
                 aes(x = long, y = lat, group = group, fill = REP.Margin)) +
    geom_polygon(data = districts_df,
                 aes(x = long, y = lat, group = group),
                 fill = NA,
                 colour = "white") +
    geom_text(data = district_centers_df,
              aes(x = long, y = lat, label = District),
              size = 5,
              colour = "white",
              show_guide = FALSE) +
    coord_equal() +
    scale_fill_gradient("Margin (Votes)",
                        limits = c(-700, 700),
                        low = "blue",
                        high = "red",
                        space = "Lab",
                        guide = "colourbar") +
    theme(axis.title = element_blank(), axis.text = element_blank()) +
    ggtitle("Allan Kittleman 2014 Margins by Precinct (Votes)")
print(g)
```

The second plot shows Allan Kittleman’s victory margins in terms of
percentage of votes in each precinct. This graph is produced
identically to the previous one, except that I use the
`Pct.REP.Margin` variable to color the precincts (instead of
`REP.Margin`) and I set the maximum red color to be used for a 70%
winning margin for Kittleman and the maximum blue color to be used for
a 70% winning margin for Courtney Watson.

```{r fig.width = 9}
g <- ggplot() +
    geom_polygon(data = precincts_df,
                 aes(x = long, y = lat, group = group, fill = Pct.REP.Margin)) +
    geom_polygon(data = districts_df,
                 aes(x = long, y = lat, group = group),
                 fill = NA,
                 colour = "white") +
    geom_text(data = district_centers_df,
              aes(x = long, y = lat, label = District),
              size = 5,
              colour = "white",
              show_guide = FALSE) +
    coord_equal() +
    scale_fill_gradient("Margin (% of Vote)",
                        limits = c(-70, 70),
                        low = "blue",
                        high = "red",
                        space = "Lab",
                        guide = "colourbar") +
    theme(axis.title = element_blank(), axis.text = element_blank()) +
    ggtitle("Allan Kittleman 2014 Margins by Precinct (% of Vote)")
print(g)
```

## Conclusion

The maps above look pretty red, not just in county council district 5,
a traditional Republican stronghold, but also in large swaths of
council districts 1 and 4. The only precincts where Courtney Watson
had truly lopsided margins of victory appear to be in the Columbia
portions of districts 2, 3, and 4.

While informative as to where Allan Kittleman had the most success
(and the least), these plots do exaggerate the size of Kittleman’s
margins of victory. That’s because visually the plot is dominated by
those precincts having the largest geographic area, which happen to be
the precincts in western Howard County in which Kittleman received his
largest margins of victory.

One way to address this issue is to draw a [cartogram][], a special
type of map in which areas are distorted so that their size is in
proportion to some underlying variable, such as (in this case) the
number of registered voters. That’s a project for the future if and
when I have time.

[cartogram]: http://en.wikipedia.org/wiki/Cartogram

## Appendix

I used the following R environment in doing the analysis for this
example:

```{r sessionInfo}
sessionInfo()
```

The underlying GDAL library for the rgdal packages is from the
KyngChaos [GDAL Complete][] distribution version 1.11 for Mac OS X.

[GDAL Complete]: http://www.kyngchaos.com/software/frameworks#gdal_complete

You can find the source code for this analysis and others at my
[HoCoData repository on GitHub][hocodata]. This document and its
source code are available for unrestricted use, distribution and
modification under the terms of the Creative Commons [CC0 1.0
Universal (CC0 1.0) Public Domain Dedication][CC0].  Stated more
simply, you’re free to do whatever you’d like with it.

[hocodata]: https://github.com/frankhecker/hocodata
[CC0]: https://creativecommons.org/publicdomain/zero/1.0/
