---
title: "Allan Kittleman’s Election-Day Victory Margins in the Howard
County 2014 General Election, Part 1"
author: "Frank Hecker"
date: "February 18, 2015"
output: html_document
---

## Introduction

This analysis uses precinct-level results from the Howard County 2014
general election (courtesy of the [Maryland State Board of
Elections][MSBoE]) to look at Allan Kittleman’s margins of victory
across the county on election day. I’m interested in the general
question of whether there seemed to be an “enthusiasm gap” in which
Kittleman’s election-day results were particularly lopsided, e.g., due
to increased turnout of Republican voters or unusually high support
for Kittleman from Democrats and unaffiliated voters.

[MSBoE]: http://elections.state.md.us/elections/2014/election_data/index.html

In this document I present the data in the form of histograms.

## Load packages

For this analysis I use the [R statistical package][] run from the
[RStudio][] development environment, along with the [dplyr][] and
[tidyr][] packages to do data manipulation and the [ggplot2][] package
to draw the histograms.

[R statistical package]: http://r-project.org
[RStudio]: http://www.rstudio.com/products/rstudio
[dplyr]: http://cran.rstudio.com/web/packages/dplyr/vignettes/introduction.html
[tidyr]: http://blog.rstudio.org/2014/07/22/introducing-tidyr/
[ggplot2]: http://ggplot2.org

```{r}
library("dplyr", warn.conflicts = FALSE)
library("tidyr")
library("ggplot2")
```

## General approach

How would one best measure relative voter enthusiasm for Allan
Kittleman vs. Courtney Watson? One measure would be how each candidate
outperformed their “expected” vote on election day, for example, how
many votes Kittleman attracted in a given precinct vs. the number of
registered Republicans in that precinct, and ditto for Watson
vis-a-vis the number of registered Democrats. A related measure would
look at Republican turnout (i.e., as a percentage of registered
Republicans) in a given precinct vs. Democratic turnout.

In this document I confine myself to looking at simple margins of
victory in each precinct. The Maryland State Board of Elections has
made available precinct-level data (in Microsoft Excel format) giving
party turnout in the 2014 general election. I’ll take a look at that
data later.

As I mentioned above, this analysis is for election day voting
only. Absentee ballots and votes cast at early voting centers are not
included in the per-precinct totals.

## Loading the data

First I download the CSV-format data containing Howard County 2014
general election results, and store a copy of the data in the local
file `Howard_By_Precinct_2014_General.csv`.

```{r}
download.file("http://elections.state.md.us/elections/2014/election_data/Howard_By_Precinct_2014_General.csv",
              "Howard_By_Precinct_2014_General.csv",
              method = "curl")
```

I then read in the CSV file for election results, and remove
extraneous spaces from the names of the offices.

```{r}
hoco_g14_df <- read.csv("Howard_By_Precinct_2014_General.csv", stringsAsFactors = FALSE)
hoco_g14_df$Office.Name <- gsub("  *$", "", hoco_g14_df$Office.Name)
str(hoco_g14_df)
```

## Data processing

I start by looking at the results for the County Executive race. I
filter the rows based on `Office.Name` and select only the columns of
interest.

```{r}
temp1_df <- hoco_g14_df %>%
    filter(Office.Name == "County Executive") %>%
    select(Election.District, Election.Precinct, Party, Election.Night.Votes)
head(temp1_df)
```

The rows where the `Party` variable has the value ‘BOT’ correspond to
votes for write-in candidates. I do a quick summarization to show that
the number of write-in votes is pretty small:

```{r}
temp1_df %>% group_by(Party) %>% summarize(Votes = sum(Election.Night.Votes))
```

To simplify the analysis I filter out the write-in votes. I also
combine the `Election.District` and `Election.Precinct` variables into
a single variable `Precinct` having the form ‘0-00’, and then discard
the original variables:

```{r}
temp2_df <- temp1_df %>%
    filter(Party != "BOT") %>%
    mutate(Precinct = paste(as.character(Election.District),
                            "-",
                            formatC(Election.Precinct, width = 2, flag = 0),
                            sep = "")) %>%
    select(-Election.District, -Election.Precinct)
head(temp2_df)
```

The problem now is that I want to compute margins of victory, and to
do that most easily I need to have both parties’ vote totals on the
same row. Enter the `spread()` function from the tidyr package. It
takes the values of the `Party` variables in different rows and
converts them into multiple column variables named after the parties
themselves. The values for the new variables are taken from the
`Election.Night.Votes` variables in the original rows for the parties.

```{r}
temp3_df <- temp2_df %>%
    spread(Party, Election.Night.Votes)
head(temp3_df)
```

The final calculation is pretty simple: I just compute a `Rep.Margin`
variable containing the absolute Republican margin of victory and a
`Pct.Rep.Margin` variable containing the Republican margin of victory
in percentage terms.

```{r}
ak_margins_df <- temp3_df %>%
    mutate(Rep.Margin = REP - DEM,
           Pct.Rep.Margin = round(100 * (REP - DEM) / (REP + DEM), 1))
head(ak_margins_df)
```

## A histogram of precincts by victory margin

Now I want to see how precincts varied in terms of the Republican
margin of victory. The easiest way to do this is using a histogram; I
first look at the percentage margins of victory:

```{r}
g <- ggplot(ak_margins_df, aes(x = Pct.Rep.Margin))
g <- g + geom_histogram()
print(g)
```

The x axis corresponds to Allan Kittleman’s different margins of
victory in various precincts, and the height of the bars corresponds
to the number of precincts for which Kittleman’s margins of victory
were in a given range.

To help better see this I change the width of the bars so they cover
ranges of 10% (0-9%, 10-19%, and so on).

```{r}
g <- ggplot(ak_margins_df, aes(x = Pct.Rep.Margin))
g <- g + geom_histogram(binwidth = 10)
g <- g + xlab("Margin of Victory (Percentage)")
g <- g + ylab("Number of Precincts")
g <- g + ggtitle("Allan Kittleman Precinct-Level Margins of Victory (%)")
print(g)
```

A couple of interesting points about this histogram: First, there were
no precincts in which Courtney Watson ran 50 or more percentage points
ahead of Allan Kittleman, and only 5 precincts in which her margin of
victory was greater than 40%. On the other hand Kittleman had what
looks like 12 precincts in which he ran more than 40% ahead of Watson.

I confirm this by sorting and filtering the `ak_margins_df` data frame
to show where Kittleman won by more than 40%:

```{r}
ak_margins_df %>%
    arrange(desc(Pct.Rep.Margin)) %>%
    filter(Pct.Rep.Margin >= 40.0)
```

and where he lost by more than 40%:

```{r}
ak_margins_df %>%
    arrange(Pct.Rep.Margin) %>%
    filter(Pct.Rep.Margin <= -40.0)
```

The other interesting thing in the histogram has to do with the
precincts in which the results were relatively close (0-10% margin
either way). It looks as if Allan Kittleman carried about twice as
many of these swing precincts as Courtney Watson.

For completeness I also show summary statistics for the percentage
margins of victory:

```{r}
summary(ak_margins_df$Pct.Rep.Margin)
```

The maximum and minimum margins of victory are as listed above.  The
median value of 4.15 means that on election day Allan Kittleman
carried half of all precincts by a 4.15% or greater margin.

Next I look at victory margins in terms of absolute numbers of votes
as opposed to percentages.

```{r}
g <- ggplot(ak_margins_df, aes(x = Rep.Margin))
g <- g + geom_histogram(binwidth = 100)
g <- g + xlab("Margin of Victory (Votes)")
g <- g + ylab("Number of Precincts")
g <- g + ggtitle("Allan Kittleman Precinct-Level Margins of Victory (Votes)")
print(g)
```

As with the other histogram this histogram is skewed to the right
(both literally and figuratively). There were no precincts in which
Courtney Watson’s election-day margin was more than 400 votes, but
several where Allan Kittleman’s election day margin of victory was
that large or larger.

## Conclusion

There are `r nrow(ak_margins_df)` precincts in Howard County.  In
voting on election day Allan Kittleman won truly lopsided victories
(40% or more winning margin) in about 10% of them, more than twice as
many as Courtney Watson. This is consistent with a greater number of
precincts that are overwhelmingly Republican (vs. Democratic), higher
Republican turnout (vs. Democratic turnout), or a decided move to
Kittleman on the part of swing voters. The evidence presented thus far
isn’t sufficient to distinguish among these possibilities.

In my next document I’ll present the same data as above, but in the
form of a precinct map.

## Appendix

I used the following R environment in doing the analysis for this
example:

```{r sessionInfo}
sessionInfo()
```

You can find the source code for this analysis and others at my
[HoCoData repository on GitHub][hocodata]. This document and its
source code are available for unrestricted use, distribution and
modification under the terms of the Creative Commons [CC0 1.0
Universal (CC0 1.0) Public Domain Dedication][CC0].  Stated more
simply, you’re free to do whatever you‘d like with it.

[hocodata]: https://github.com/frankhecker/hocodata
[CC0]: https://creativecommons.org/publicdomain/zero/1.0/
